---
title: "Soil contamination maps by kriging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soil contamination maps by kriging}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Why this article?

Every spacemodR simulation starts from a **raster of soil contamination**
(e.g. `ground_concentration_cd_compressed.tif` used in all the other articles).
In the field, however, soil is only sampled at a few hundred **points**.
This article explains, step by step, how to go from these points to a
continuous raster, and which interpolation method gives the most reliable
map for the Metaleurop site.

The workflow has 8 steps:

| Step | What we do | spacemodR function |
|:----:|:-----------|:-------------------|
| 1 | Look at the data and choose the scale (log10) | `data(soil_metaleurop)` |
| 2 | Understand the large-scale trend (distance to the smelter) | — |
| 3 | Build the output grid and the covariate rasters | `soil_grid()`, `source_covariates()` |
| 4 | Compute the empirical variogram | `soil_variogram()` |
| 5 | Fit a variogram model | `fit_soil_variogram()` |
| 6 | Compare interpolation methods by cross-validation | `cv_soil()`, `kernel_smooth_soil()` |
| 7 | Produce the final Cd, Pb and Zn rasters (+ uncertainty) | `krige_soil()` |
| 8 | Check the maps and plug them into spacemodR | `load_raster_extdata()` |

**Short answer** (justified in step 6): for the three metals, the best method
is **kriging with external drift (KED)** of `log10(concentration)`, using
`log10(distance to the smelter)` as drift and an **exponential** variogram.

```{r setup}
library(spacemodR)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
library(patchwork)
```

# Step 1 — The data

The data set `soil_metaleurop` contains `r nrow(soil_metaleurop)` topsoil
samples collected around the former Metaleurop Nord lead and zinc smelter
(northern France), with total Cd, Pb and Zn concentrations (mg/kg).
The coordinates are in Lambert-93 (EPSG:2154), **in metres**: this is required,
since kriging works with distances.

```{r data}
data(soil_metaleurop)
soil_metaleurop
summary(sf::st_drop_geometry(soil_metaleurop)[, c("cd", "pb", "zn")])
```

The concentrations are extremely **skewed**: for cadmium the median is around
5 mg/kg but the maximum is above 2000 mg/kg. Kriging is a linear method
(a weighted average of neighbouring values) and is very sensitive to such
extreme values. On a **log10 scale**, the distributions become nearly
symmetric (log-normal data), so all the interpolation is done on
`log10(concentration)`. This is the default (`log10 = TRUE`) in all spacemodR
functions.

```{r histograms, fig.height = 4}
long <- data.frame(
  metal = rep(c("Cd", "Pb", "Zn"), each = nrow(soil_metaleurop)),
  value = c(soil_metaleurop$cd, soil_metaleurop$pb, soil_metaleurop$zn)
)
p_raw <- ggplot(long, aes(value)) +
  geom_histogram(bins = 40, fill = "grey40") +
  facet_wrap(~metal, scales = "free") +
  labs(title = "Raw scale (mg/kg)", x = NULL) + theme_minimal()
p_log <- ggplot(long, aes(log10(value))) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~metal, scales = "free") +
  labs(title = "log10 scale", x = NULL) + theme_minimal()
p_raw / p_log
```

The three metals are highly correlated (same emission source), so the same
method should suit all of them:

```{r correlation}
cor(log10(sf::st_drop_geometry(soil_metaleurop)[, c("cd", "pb", "zn")]))
```

# Step 2 — The large-scale trend: distance to the smelter

Contamination comes from the atmospheric fallout of a **single point source**:
the smelter chimney. The data set gives the distance of each sample to the
smelter (`dist`), but not the smelter location itself. We recover it by
looking for the point whose distance to every sample best matches `dist`
(a simple least-squares triangulation):

```{r smelter}
xy <- sf::st_coordinates(soil_metaleurop)
misfit <- function(p) {
  sum((sqrt((xy[, 1] - p[1])^2 + (xy[, 2] - p[2])^2) - soil_metaleurop$dist)^2)
}
fit <- optim(colMeans(xy), misfit)
smelter <- sf::st_sfc(sf::st_point(fit$par), crs = 2154)
smelter
# average mismatch with the `dist` column (m):
sqrt(fit$value / nrow(xy))
```

The match is almost perfect (about 1 m), and the point lies at
`r paste(round(sf::st_coordinates(sf::st_transform(smelter, 4326)), 4), collapse = "° E, ")`° N,
i.e. on the former Metaleurop site at Noyelles-Godault.

```{r map_points, fig.height = 6}
ggplot() +
  geom_sf(data = soil_metaleurop, aes(colour = log10(cd)), size = 1.8) +
  geom_sf(data = smelter, shape = 8, size = 5, colour = "red", stroke = 1.5) +
  scale_colour_viridis_c(name = "log10 Cd\n(mg/kg)") +
  labs(title = "Sampling points (red star: smelter)") +
  theme_minimal()
```

Plotting the concentrations against the distance on a log-log scale reveals a
strong **linear trend**: concentrations decrease as a power law of the
distance, as expected for atmospheric deposition.

```{r distance_trend, fig.height = 4}
long$dist <- rep(soil_metaleurop$dist, 3)
ggplot(long, aes(log10(dist), log10(value))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", colour = "red") +
  facet_wrap(~metal, scales = "free_y") +
  labs(x = "log10(distance to smelter, m)", y = "log10(concentration)") +
  theme_minimal()

summary(lm(log10(cd) ~ log10(dist), data = soil_metaleurop))$r.squared
```

The distance alone explains about **half of the variance** of log10(Cd). This
is a precious information for interpolation, especially **far from the
samples**: ordinary kriging would there predict the global mean of the data,
whereas a model including the distance keeps decreasing away from the
smelter. This motivates testing *kriging with external drift* below.

# Step 3 — Output grid and covariate rasters

## 3a. The grid

`soil_grid()` builds an empty raster around the points, with a given
resolution `res` and a `margin` around the bounding box of the samples.
Here: 25 m cells and 1 km of margin, as for the rasters already shipped
with the package.

```{r grid}
grid <- soil_grid(soil_metaleurop, res = 25, margin = 1000)
grid
```

## 3b. The covariates

`source_covariates()` computes, for every cell of the grid:

* `log10_dist`: the log10 of the distance to the source (the drift found in
  step 2);
* `wind` (optional): the frequency of the wind sector in which the cell lies,
  seen from the source. The data set gives this frequency (`wind`) for each
  sample in 20° sectors, so we rebuild the wind rose from it:

```{r wind_rose}
sector <- (round(soil_metaleurop$angle / 20) %% 18) * 20
wind_rose <- aggregate(
  soil_metaleurop$wind, by = list(direction = sector),
  FUN = function(x) as.numeric(names(which.max(table(x))))
)
names(wind_rose)[2] <- "freq"
wind_rose
```

```{r covariates, fig.height = 4}
covariates <- source_covariates(grid, smelter, wind_rose = wind_rose)
covariates

plot_cov <- function(layer, title) {
  ggplot() +
    geom_spatraster(data = covariates[[layer]]) +
    scale_fill_viridis_c(name = NULL) +
    geom_sf(data = smelter, shape = 8, size = 3, colour = "red") +
    labs(title = title) + theme_minimal()
}
plot_cov("log10_dist", "log10(distance to the smelter, m)") +
  plot_cov("wind", "Wind frequency of the sector (%)")
```

The wind layer shows that the north-east of the smelter is the most exposed,
downwind of the dominant south-westerly winds.

# Step 4 — The empirical variogram

Kriging relies on one idea: **two close samples are more alike than two
distant samples**. The *semi-variogram* quantifies this: for all pairs of
points separated by a distance $h$ (grouped in distance classes, or *lags*),

$$\gamma(h) = \frac{1}{2 N(h)} \sum_{(i,j) \,:\, d_{ij} \approx h} \left(z_i - z_j\right)^2$$

where $z$ is the log10 concentration and $N(h)$ the number of pairs.
A variogram that increases with $h$ means there is spatial structure that
kriging can exploit.

`soil_variogram()` computes it. By default it uses distances up to one third
of the study area diagonal (`cutoff`) in 15 classes (`width`).

```{r variogram}
v_ok <- soil_variogram(soil_metaleurop, "cd")
head(v_ok)
```

When a drift is used (KED), the spatial structure must be described on the
**residuals of the trend**, not on the raw values; this is done by passing the
covariates:

```{r variogram_ked}
v_ked <- soil_variogram(soil_metaleurop, "cd",
                        covariates = covariates[["log10_dist"]])
```

```{r variogram_plot, fig.height = 4}
both <- rbind(
  data.frame(as.data.frame(unclass(v_ok)), type = "raw values (OK)"),
  data.frame(as.data.frame(unclass(v_ked)), type = "residuals of log10(dist) (KED)")
)
ggplot(both, aes(dist, gamma, colour = type)) +
  geom_point(aes(size = np)) + geom_line() +
  expand_limits(y = 0) +
  labs(x = "Distance h (m)", y = expression(gamma(h)),
       colour = NULL, size = "Number of pairs",
       title = "Empirical variograms of log10(Cd)") +
  theme_minimal()
```

Reading the plot:

* the **raw** variogram keeps increasing over several kilometres: this is the
  signature of the large-scale trend (values decrease with the distance to
  the smelter);
* once the trend is removed, the **residual** variogram is about half as
  high and almost flat beyond 1–1.5 km: the remaining spatial dependence is
  short-range, the long-range part being explained by the distance;
* neither curve starts at 0: the value at the origin is the **nugget**,
  i.e. variability between very close samples (measurement error, very
  local heterogeneity, the extreme outlier of the data set...).

# Step 5 — Fitting a variogram model

Kriging needs a continuous function $\gamma(h)$, so a model is fitted to the
empirical points. Every model has three parameters:

* **nugget** $c_0$: the jump at the origin;
* **partial sill** $c$: the additional variance reached at large distances
  (total sill = nugget + partial sill);
* **range** $a$: how fast the sill is reached (the distance at which samples
  stop being correlated is about $3a$ for the exponential model, $a$ for the
  spherical one).

Four models are available: exponential (`"Exp"`), spherical (`"Sph"`),
Gaussian (`"Gau"`) and Matérn (`"Mat"`). `fit_soil_variogram()` fits them by
weighted least squares (more weight to short distances and to classes with
many pairs) and returns the one with the lowest error, together with the
comparison table.

```{r fit_variogram}
m_ok <- fit_soil_variogram(v_ok, model = c("Exp", "Sph", "Gau"))
m_ok
m_ok$comparison

m_ked <- fit_soil_variogram(v_ked, model = c("Exp", "Sph", "Gau"))
m_ked
```

```{r fit_plot, fig.height = 4}
(plot(m_ok) + ggtitle("OK: raw log10(Cd)")) +
  (plot(m_ked) + ggtitle("KED: residuals of the distance trend")) +
  plot_layout(guides = "collect")
```

The models are very close to each other. **The best fit to the variogram
points does not necessarily give the best map**: this is what the old Python
workflow used (ranking models by the $R^2$ of the variogram fit). A more
reliable criterion is the quality of the *predictions*, assessed by
cross-validation in the next step.

# Step 6 — Which method gives the best map? Cross-validation

## 6a. Principle

In **k-fold cross-validation** (`cv_soil()`), the points are split into 10
groups. Each group is removed in turn, the whole method (including the
variogram fit) is run on the 9 other groups, and the removed points are
predicted. Comparing predictions with observations gives:

* **RMSE** (root mean squared error, log10 units) — the main criterion,
  lower is better;
* **MAE** (mean absolute error) — less sensitive to the outlier;
* **bias** (mean error) — should be close to 0;
* **R2** — share of the variance of the data reproduced by the predictions.

Two ways of making the groups are used:

* **random** folds: the removed points usually have close neighbours left,
  this measures the quality of the interpolation *between* samples;
* **spatial** folds (`block_size = 1500`): whole 1.5 km × 1.5 km blocks are
  removed, which mimics the prediction of areas **without any sample nearby**
  (e.g. the margin of the map). The block size is of the order of the
  distance at which the residual variogram flattens.

## 6b. The candidates

1. **Kernel smoothing** (`kernel_smooth_soil()`): a Gaussian weighted mean of
   the samples, built with the dispersal kernel of the package
   (`compute_kernel()`) and a convolution (`terra::focal()`). Simple, but it
   ignores the spatial structure and its smoothing radius must be chosen.
2. **Ordinary kriging (OK)** with an exponential, spherical or Gaussian
   variogram: what the old Python workflow (`pykrige`) did.
3. **Kriging with external drift (KED)** on `log10_dist`: a linear trend in
   log-distance to the smelter + kriging of the residuals.
4. **KED on `log10_dist` + `wind`**.

```{r cv_setup}
candidates <- list(
  "Kernel (radius 300 m)"  = list(method = "kernel", radius = 300),
  "OK - Exp"               = list(model = "Exp"),
  "OK - Sph"               = list(model = "Sph"),
  "OK - Gau"               = list(model = "Gau"),
  "KED dist - Exp"         = list(model = "Exp", covariates = covariates[["log10_dist"]]),
  "KED dist + wind - Exp"  = list(model = "Exp", covariates = covariates)
)

run_cv <- function(metal, block_size = NULL) {
  do.call(rbind, lapply(names(candidates), function(nm) {
    args <- c(list(points = soil_metaleurop, value = metal,
                   block_size = block_size), candidates[[nm]])
    cbind(method = nm, summary(do.call(cv_soil, args)))
  }))
}
```

```{r cv_run, cache = TRUE, cache.extra = packageVersion("spacemodR")}
cv_results <- do.call(rbind, lapply(c("cd", "pb", "zn"), function(metal) {
  rbind(cbind(metal = metal, folds = "random",  run_cv(metal)),
        cbind(metal = metal, folds = "spatial", run_cv(metal, block_size = 1500)))
}))
cv_results$method <- factor(cv_results$method, levels = names(candidates))
```

```{r cv_table}
knitr::kable(
  cv_results[order(cv_results$folds, cv_results$metal, cv_results$RMSE),
             c("folds", "metal", "method", "RMSE", "MAE", "bias", "R2", "n")],
  digits = 3, row.names = FALSE
)
```

```{r cv_plot, fig.height = 4.5}
ggplot(cv_results, aes(x = RMSE, y = method, colour = folds)) +
  geom_point(size = 3) +
  facet_wrap(~metal, scales = "free_x") +
  scale_y_discrete(limits = rev) +
  labs(x = "Cross-validation RMSE (log10 units, lower is better)", y = NULL,
       colour = "Folds") +
  theme_minimal()
```

## 6c. Conclusion of the comparison

* **Kernel smoothing** is competitive with random folds (it behaves like a
  smooth ordinary kriging), but it is clearly the worst with **spatial
  folds**: far from the samples, it simply averages whatever lies within its
  window, without any model of how values evolve with distance.
* The **variogram model of ordinary kriging** (Exp / Sph / Gau) changes the
  error only marginally.
* **KED with the distance to the smelter** gives the lowest error for the
  three metals, and the gain is largest with **spatial folds**: the drift is
  most useful where samples are scarce.
* Adding the **wind** does not bring a consistent improvement (once the
  distance is known, the wind frequency does not help predicting, and an
  extra parameter adds noise). It is not retained.

We also tested, outside this article, a generalized additive model
(`mgcv::gam(log10(cd) ~ log10(dist) + wind + s(x, y))`) and a random forest
(`ranger`) on coordinates, distance and wind: neither beat KED, so no
additional package dependency is needed.

**Retained method: KED, drift = log10(distance to the smelter), exponential
variogram, on log10 concentrations.**

The exponential model is kept (rather than the one with the best variogram fit)
because it is the most robust across metals in spatial cross-validation, and
it is the classical choice for contaminated soils (continuous but rough
surfaces).

# Step 7 — Final rasters of Cd, Pb and Zn

`krige_soil()` does steps 4, 5 and the prediction in one call. To obtain maps
that can directly replace the rasters already used in spacemodR, we use the
existing raster as `template` (same extent and resolution).

```{r final_template}
template <- load_raster_extdata("ground_concentration_cd_compressed.tif")
dist_template <- source_covariates(template, smelter)   # layer "log10_dist"
```

```{r final_krige, message = TRUE, cache = TRUE, cache.extra = packageVersion("spacemodR"), echo = 1:7}
soil_maps <- lapply(c("cd", "pb", "zn"), function(metal) {
  krige_soil(soil_metaleurop, metal,
             template = template,
             covariates = dist_template,
             model = "Exp")
})
names(soil_maps) <- c("cd", "pb", "zn")
soil_maps_packed <- lapply(soil_maps, terra::wrap)
```

```{r final_krige_unpack, include = FALSE}
# terra rasters are pointers to C++ objects and cannot be stored in the knitr
# cache: they are packed (terra::wrap) at the end of the cached chunk above
# and restored here.
soil_maps <- lapply(soil_maps_packed, terra::unwrap)
```

```{r final_krige_show}
soil_maps$cd
```

Note the messages about **clamping**: the closest sample is about
`r round(min(soil_metaleurop$dist))` m from the chimney, so the
`log10_dist` values of the cells closer than that are outside the range of
the data. Extrapolating the linear trend there would predict unrealistic
values (several thousand mg/kg of Cd inside the factory site). By default
(`clamp = TRUE`), `krige_soil()` therefore restricts each covariate to the
range observed at the samples: within ~`r round(min(soil_metaleurop$dist))` m
of the chimney, and beyond the farthest sample, the trend is kept constant.

Each result has three layers:

* `cd_log10`: the prediction on the log10 scale;
* `cd_var`: the **kriging variance** (log10 scale), an estimate of the
  uncertainty of each cell;
* `cd`: the back-transformed concentration $10^{\text{prediction}}$ in mg/kg.
  Note that it is the **median** concentration expected in the cell (the mean
  of a log-normal variable is higher: $10^{\mu + \ln(10)\,\sigma^2/2}$).

```{r final_maps, fig.height = 4}
plot_map <- function(metal) {
  ggplot() +
    geom_spatraster(data = soil_maps[[metal]][[paste0(metal, "_log10")]]) +
    scale_fill_viridis_c(name = "log10\n(mg/kg)") +
    geom_sf(data = smelter, shape = 8, size = 2, colour = "red") +
    labs(title = toupper(metal)) +
    theme_minimal() +
    theme(axis.text = element_blank(), legend.position = "bottom")
}
plot_map("cd") + plot_map("pb") + plot_map("zn") +
  plot_annotation(title = "KED predictions (log10 concentration)")
```

## Uncertainty

The kriging standard deviation ($\sqrt{\text{variance}}$) is low close to the
samples and grows in the margin, where no data are available. It is a
useful map to know where the contamination layer, and therefore the whole
spacemodR risk assessment, is less reliable.

```{r uncertainty, fig.height = 5}
ggplot() +
  geom_spatraster(data = sqrt(soil_maps$cd[["cd_var"]])) +
  geom_sf(data = soil_metaleurop, size = 0.3, colour = "white") +
  scale_fill_viridis_c(name = "sd\n(log10)", option = "magma") +
  labs(title = "Kriging standard deviation of log10(Cd)") +
  theme_minimal()
```

A rough 95% interval of the concentration in each cell is
$10^{\text{pred} \pm 1.96\,\text{sd}}$:

```{r interval, fig.height = 4}
sd_cd <- sqrt(soil_maps$cd[["cd_var"]])
interval <- c(10^(soil_maps$cd[["cd_log10"]] - 1.96 * sd_cd),
              soil_maps$cd[["cd"]],
              10^(soil_maps$cd[["cd_log10"]] + 1.96 * sd_cd))
names(interval) <- c("lower 2.5%", "median", "upper 97.5%")
ggplot() +
  geom_spatraster(data = log10(interval)) +
  facet_wrap(~lyr) +
  scale_fill_viridis_c(name = "log10 Cd\n(mg/kg)") +
  theme_minimal() + theme(axis.text = element_blank())
```

# Step 8 — Checks and use in spacemodR

## 8a. Behaviour far from the samples

The main practical difference between OK and KED lies in the margin of the map.
We extract both predictions along a transect starting at the smelter and going
to the south-west corner of the map:

```{r ok_krige, cache = TRUE, cache.extra = packageVersion("spacemodR"), echo = 1:2}
ok_cd <- krige_soil(soil_metaleurop, "cd", template = template,
                    model = "Exp", verbose = FALSE)
ok_cd_packed <- terra::wrap(ok_cd)
```

```{r ok_krige_unpack, include = FALSE}
ok_cd <- terra::unwrap(ok_cd_packed)
```

```{r transect, fig.height = 4}

corner <- c(xmin(template), ymin(template))
start <- sf::st_coordinates(smelter)[1, ]
s <- seq(0, 1, length.out = 300)
transect <- cbind(x = start[1] + s * (corner[1] - start[1]),
                  y = start[2] + s * (corner[2] - start[2]))
dist_km <- sqrt((transect[, 1] - start[1])^2 + (transect[, 2] - start[2])^2) / 1000

# last distance with a sample closer than 300 m from the transect
pts_xy <- sf::st_coordinates(soil_metaleurop)
near <- apply(transect, 1, function(p) min(sqrt((pts_xy[, 1] - p[1])^2 + (pts_xy[, 2] - p[2])^2)))
last_sample <- max(dist_km[near < 300])

profile <- rbind(
  data.frame(dist_km, value = extract(ok_cd[["cd_log10"]], transect)[, 1], method = "OK"),
  data.frame(dist_km, value = extract(soil_maps$cd[["cd_log10"]], transect)[, 1], method = "KED (distance)")
)
ggplot(profile, aes(dist_km, value, colour = method)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = last_sample, linetype = 2) +
  annotate("text", x = last_sample, y = max(profile$value), hjust = -0.05,
           label = "no more samples beyond this line") +
  labs(x = "Distance from the smelter along the transect (km)",
       y = "Predicted log10(Cd)", colour = NULL) +
  theme_minimal()
```

Beyond the last samples, OK goes back to the **mean of the data**, which is
dominated by the many samples taken close to the smelter: it overestimates the
contamination far away. KED keeps following the physically sensible decrease
with distance.

## 8b. Comparison with the previous rasters

The rasters shipped in `inst/extdata` were produced by the old Python
workflow (ordinary kriging). The difference with the new KED maps:

```{r compare_old, fig.height = 4}
old_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
diff_cd <- soil_maps$cd[["cd_log10"]] - old_cd
names(diff_cd) <- "KED - old OK"

lim <- max(abs(values(diff_cd)), na.rm = TRUE)
ggplot() +
  geom_spatraster(data = diff_cd) +
  scale_fill_distiller(palette = "RdBu", limits = c(-lim, lim),
                       name = "difference\n(log10)") +
  geom_sf(data = soil_metaleurop, size = 0.2) +
  labs(title = "log10(Cd): new KED map minus previous map") +
  theme_minimal()
```

```{r compare_old_stats}
# Inside / outside the convex hull of the samples
hull <- terra::vect(sf::st_convex_hull(sf::st_union(soil_metaleurop)))
inside  <- values(terra::mask(diff_cd, hull), mat = FALSE)
outside <- values(terra::mask(diff_cd, hull, inverse = TRUE), mat = FALSE)
rbind(inside  = quantile(inside,  c(0.05, 0.5, 0.95), na.rm = TRUE),
      outside = quantile(outside, c(0.05, 0.5, 0.95), na.rm = TRUE))
```

Inside the sampled area the two maps are close (differences are local and
mostly small, they come from the different variogram and from the drift).
Outside the sampled area, the new map is systematically **lower**, because
OK reverted to the mean of the data far from the samples (see 8a).

## 8c. Use in a spacemodR workflow

The resulting layer behaves exactly like the rasters used in the other
articles: it can be saved, masked to a region of interest, and used as the
soil layer of a `spacemodel`.

```{r use, eval = FALSE}
# Save the log10 Cd map
terra::writeRaster(soil_maps$cd[["cd_log10"]], "ground_concentration_cd.tif",
                   overwrite = TRUE)

# Or use it directly, e.g. restricted to the region of interest
data(roi_metaleurop)
ground_cd <- terra::mask(soil_maps$cd[["cd_log10"]],
                         terra::vect(sf::st_transform(roi_metaleurop, 2154)))
```

# Summary: the whole workflow in a few lines

```{r summary, eval = FALSE}
library(spacemodR)
data(soil_metaleurop)

# 1. Source location and covariate raster
smelter <- sf::st_sfc(sf::st_point(c(701156, 7036799)), crs = 2154)
grid <- soil_grid(soil_metaleurop, res = 25, margin = 1000)
dist_cov <- source_covariates(grid, smelter)          # layer "log10_dist"

# 2. (optional) inspect the variogram
v <- soil_variogram(soil_metaleurop, "cd", covariates = dist_cov)
plot(fit_soil_variogram(v, model = c("Exp", "Sph", "Gau")))

# 3. (optional) check the prediction error
summary(cv_soil(soil_metaleurop, "cd", covariates = dist_cov, model = "Exp"))

# 4. Kriging with external drift
cd_map <- krige_soil(soil_metaleurop, "cd", covariates = dist_cov, model = "Exp")
terra::plot(cd_map)
```

# Appendix — Implementation notes

* The kriging is implemented directly in spacemodR (packages `sf`, `terra` and
  `stats` only): with a few hundred samples, the kriging system can be solved
  globally (all samples are used for each prediction). It follows the
  conventions of the reference package **gstat** (same variogram models and
  parameters, same weighted least-squares fit), and gives the same predictions
  and variances as `gstat::krige()` (checked to $10^{-7}$) for both ordinary
  kriging and kriging with external drift.
* Duplicated sampling locations are averaged (otherwise the kriging system is
  singular): this is why the cross-validation tables report `n` = 587
  instead of 595.
* For data sets with strong outliers, `soil_variogram(..., estimator = "robust")`
  uses the Cressie-Hawkins estimator, which is less sensitive to them.
* Kernel smoothing reuses `compute_kernel()`, the Gaussian kernel used for
  dispersal in spacemodR.

```{r gstat_check, eval = requireNamespace("gstat", quietly = TRUE)}
# Optional check against gstat (only run if gstat is installed)
pts <- soil_metaleurop[!duplicated(sf::st_coordinates(soil_metaleurop)), ]
pts$cd_log10 <- log10(pts$cd)
m <- soil_vgm("Exp", psill = 0.17, range = 2000, nugget = 0.04)
g <- gstat::vgm(psill = 0.17, model = "Exp", range = 2000, nugget = 0.04)

small_grid <- soil_grid(soil_metaleurop, res = 500, margin = 0)
ours <- krige_soil(pts, "cd", template = small_grid, vgm = m, verbose = FALSE)
cells <- sf::st_as_sf(as.data.frame(xyFromCell(small_grid, 1:ncell(small_grid))),
                      coords = c("x", "y"), crs = 2154)
ref <- gstat::krige(cd_log10 ~ 1, pts, cells, model = g, debug.level = 0)
max(abs(ref$var1.pred - values(ours[["cd_log10"]])))
```

# References

* Pebesma, E. J. (2004). Multivariable geostatistics in S: the gstat package.
  *Computers & Geosciences*, 30(7), 683–691.
* Hengl, T., Heuvelink, G. B. M., & Rossiter, D. G. (2007). About
  regression-kriging: from equations to case studies.
  *Computers & Geosciences*, 33(10), 1301–1315.
* Webster, R., & Oliver, M. A. (2007). *Geostatistics for Environmental
  Scientists* (2nd ed.). Wiley.
