---
title: "Food Intake Rate"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Food_Intake}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=10,
  fig.height=5,
  warning = FALSE,
  message = FALSE
)
library(ggplot2)
library(spacemodR)

load_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(readRDS(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
```


The calculation of Daily Food Intake (DFI), expressed in grams of fresh matter per day ($g/day$), serves as the central pivot for translating physiological energy expenditure into actual predation or herbivory pressures. This estimation process requires the alignment of four critical variables: the Field Metabolic Rate (FMR), nutrient assimilation efficiency, the energy density of the consumed items, and their water content.

# Energetic Needs

Estimating food consumption from metabolism relies on an energy flux model where the maximum intake ($I$) can be derived from the following formula:

$$
I = \frac{FMR}{e_{a}(1 - e_{p})i}
$$

In this equation, the variables are defined as follows:

- $FMR$ represents the field metabolic rate ($kJ/day$).
- $e_{a}$ is the assimilation efficiency (ranging from $0$ to $1$). 
- $e_{p}$ is the production efficiency. 
- $i$ corresponds to the caloric content or energy density of the food item ($kJ/g$).

Sometimes, the equation is written:


$$
FIR = \frac{DEE}{FE \times (1 - MC)*AE} = \frac{FMR}{i \times (1 - MC)*e_{a}}
$$

If the animal is an adult with a stable body mass, its production efficiency
($e_{p}$) is close to 0.

## The FmrBT Database

The database provided by @decastro2025fmrbt, named `FmrBT`, constitutes a
comprehensive resource of Field Metabolism Rates ($FMR$), body mass, and ambient
temperature, covering over 700 identified species.

Unlike the Basal Metabolic Rate (BMR), the $FMR$ captures energy expended 
within a natural habitat, including locomotion, foraging, digestion, 
and thermoregulation. This database is crucial because it explicitly 
integrates the effects of temperature, a factor that accelerates metabolic
chemical reactions, particularly in ectotherms.

- FMR $[kJ/d]$: Field metabolic rate per individual ($kJ/day$).
- Mass $[g]$: Body mass of the individual ($grams$).
- Temp $[Celsius]$: The temperature recorded during the study ($Celsius$).
- Endotherm $[\text{binary}]$: The thermal status of the species, where $1$ 
represents an endotherm (organism that maintains its body at a metabolically 
favorable temperature) and $0$ represents an ectotherm (internal physiological 
sources of heat quite negligible).


```{r FmrBT_DB}
# FmrBT <- read.csv("raw_data/FieldMR_SciDataPub.csv")
data("FmrBT")

FmrBT$Endotherm = ifelse(FmrBT$Endotherm==0, FALSE, TRUE)
head(FmrBT)
```

There is a log-log relationship between Mass and Field Metabolism.

```{r}
ggplot(data=FmrBT) +
  theme_minimal() +
  labs(
    title="Log-Log Relation between Mass and Field Metabolism Rates",
    x="Mass [g], log10-scale", y="Field Metabolism Rate [kJ/day], log10"
    ) +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual("Endotherm", values=c("#1144aa", "#aa4411")) +
  geom_point(aes(x=Mass_g, y=FMR_kJ_d, color=Endotherm), alpha=0.8)

ggplot(data=FmrBT[FmrBT$Endotherm, ]) +
  theme_minimal() +
  labs(
    title="Log-Log Relation between Mass and FMR (only Endotherm)",
    x="Mass [g], log10-scale", y="Field Metabolism Rate [kJ/day], log10"
    ) +
  scale_x_log10() + scale_y_log10() +
  scale_shape_manual(values = c(1,16)) +
  geom_point(aes(x=Mass_g, y=FMR_kJ_d, color=Order, shape=Class),  alpha=0.8)
```


```{r}
# FmrBT_Endotherm_NAMES <- read.csv("raw_data/FieldMR_SciDataPub_Endotherm_NAMES.csv")
# FmrBT_Endotherm_NAMES
```

# Food Item Proportion


Apply the dietary category percentages from the EltonTraits 1.0 database (for birds and mammals) or MammalBase.

For reptiles, refer to data provided by the GARD Initiative.

Partitionnement du Régime Alimentaire : Appliquer les pourcentages de catégories
de la base EltonTraits 1.0 (oiseaux/mammifères) ou de MammalBase.

Pour les reptiles, se référer aux données de l'initiative GARD

EltonTraits : 

=> EXPLIQUER CE QU'ON A DANS LES COLONNES !!!
```{r}
data("DBFunc_BirdFuncDat")
data("DBFunc_MamFuncDat")

# head(DBFunc_MamFuncDat)
```

## Energy Density in Items 

Assignation de la Densité Énergétique ($kJ/g$ MS) :

Végétaux : Utiliser les valeurs LCV de la base TRY.

Invertébrés : Appliquer le modèle AFDW ($log_{10}(ED) = 1,07 \times log_{10}(AFDW) - 0,80$) ou les constantes de Cummins & Wuycheck.

Vertébrés : Utiliser le paramètre $\xi_{WE}$ de Add-my-pet.

Une avancée significative dans l'estimation de la densité énergétique repose sur l'utilisation du pourcentage d'AFDW. Le poids sec total ($DW$) inclut souvent des composants inorganiques non digestibles comme les os, les coquilles ou les sels, qui varient considérablement entre les organismes et peuvent fausser les estimations énergétiques. En isolant la fraction organique, le modèle AFDW permet d'estimer la densité énergétique des invertébrés, des vertébrés et des plantes avec une seule équation.L'équation de régression globale identifiée est :$$\log_{10}(ED) = 1,07 \times \log_{10}(AFDW) - 0,80$$

## Dry/Fresh Weight

Conversion en Matière Fraîche ($g/jour$) : Utiliser les ratios de teneur en eau par taxon (ex: diviser la masse sèche par 0,22 pour les invertébrés ou 0,30 pour les mammifères).

## Assimilation rate

Correction de l'Assimilation : Appliquer le taux d'assimilation ($e_{a}$) spécifique à l'item alimentaire (ex: 0,7 pour les herbivores, 0,8 pour les carnivores).
