---
title: "Habitat Layer from EUNIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Habitat Layer from EUNIS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The [Habitat Layer](core_Habitat.html) guide builds species habitat maps from
the French OCS-GE land cover and the `ocsge_species_dict` dataset. This guide
does the same from the **EUNIS** classification (European Nature Information
System), using the `eunis_species_dict` dataset.

EUNIS is a useful alternative when:

* the study area is **outside France**: the EEA *Ecosystem Type Map v3.1*
  covers the 39 EEA countries;
* the species depends on habitats that OCS-GE cannot separate: OCS-GE describes
  land *cover* only, so reedbeds, wet meadows, bogs, dunes or arable land are
  merged into broad classes. EUNIS Level 2 has 45 terrestrial classes,
  including 6 wetland types (D1–D6), 6 grassland types (E1–E7) and 10 scrub
  types (F1–FB).

The trade-off is resolution: the EUNIS map is 100 m, while OCS-GE is vector
data mapped at about 1:5,000 scale.

We again focus on the Wood mouse (*Apodemus sylvaticus*), then compare a few
other species.

```{r setup}
library(spacemodR)
library(dplyr)
library(ggplot2)
library(terra)
library(tidyterra)
library(leaflet)
library(patchwork)
```

# The database

## How it was built

`eunis_species_dict` covers the same 935 European vertebrates as
`ocsge_species_dict`, with the same EEA habitat descriptions as input. Each
species was scored against the 45 EUNIS Level 2 classes, each described by a
short ecological summary (e.g. *G5: "Lines of trees, small woodlands, recently
felled woodland: edge-rich, fragmented"*).

Compared with `ocsge_species_dict`, the scoring was changed to make the values
easier to compare across species and directly usable in connectivity models:

* **Anchored scale.** Every score uses the same reference points
  (0 unused or avoided, 2 marginal, 5 secondary, 8 main, 10 core habitat),
  so that a 5 means the same thing for a toad and for an eagle.
* **Breeding separated from foraging.** Many species breed in one habitat and
  feed in another (amphibians in ponds and woods, raptors on cliffs and in
  steppe).
* **Taxon-aware movement.** Flight, swimming and burrowing are taken into
  account, so a road is a strong barrier for a toad but not for a kite.
* **Resistance derived, not scored.** Resistance is computed from permeability
  on the 1–100 log scale expected by Omniscape, so the two can never contradict
  each other and no further transformation is needed.
* **Species-level flags** for confidence in the source description and
  presence in France.

Scores were assigned by a large language model (Claude Opus 5.5) reading the
descriptions; they are expert-judgement values, not fitted to occurrence data.

## What each column means

```{r load_dict}
data("eunis_species_dict")
data("ref_eunis")
str(eunis_species_dict)
```

| Column | Level | Range | Meaning |
|---|---|---|---|
| `code_eunis` | class | – | EUNIS Level 2 code, matches `ref_eunis$code` (with `level == "L2"`) and the raster returned by `get_eunis_data(level = "L2")`. |
| `nom_espece` | species | – | Scientific name, `Genus_species`. |
| `sppID` | species | – | EEA identifier; first letter gives the group (A, R, B, M). |
| `taxon_group` | species | – | amphibian, reptile, bird or mammal. |
| `suitability` | class | 0–10 | Overall habitat quality. **Use as the source / habitat weight.** |
| `weight_breeding` | class | 0–10 | Value as breeding, nesting, denning, roosting or larval site. |
| `weight_foraging` | class | 0–10 | Value for feeding. |
| `permeability` | class | 0–10 | Ease of movement (10 free, 7 readily, 4 reluctantly, 1 strong barrier, 0 impassable). |
| `resistance` | class | 1–100 | `10^(2 * (1 - permeability / 10))`: 10 → 1, 5 → 10, 0 → 100. **Use as the resistance layer.** |
| `habitat_breadth` | species | 0–45 | Number of classes with `suitability >= 5`. |
| `specialization` | species | 0–1 | 1 − normalised Shannon evenness of the suitability profile: 0 generalist, 1 specialist. |
| `confidence` | species | 1–3 | Quality of the source description: 1 vague or empty, 2 medium, 3 detailed. |
| `in_france` | species | 0/1 | 1 if the species occurs natively and regularly in metropolitan France. |

The relation between permeability and resistance:

```{r perm_res, fig.height=3.5, fig.width=6}
data.frame(permeability = 0:10) %>%
  mutate(resistance = 10^(2 * (1 - permeability / 10))) %>%
  ggplot(aes(permeability, resistance)) +
  geom_line(colour = "#2a78d6", linewidth = 1) +
  geom_point(colour = "#2a78d6", size = 2) +
  scale_y_log10(breaks = c(1, 2.5, 10, 25, 100)) +
  scale_x_continuous(breaks = 0:10) +
  theme_minimal() +
  labs(x = "Permeability (0–10)", y = "Resistance (log scale)")
```

Species-level columns help select or weigh species before mapping, for example
the specialists present in France with a reliable description:

```{r species_level}
eunis_species_dict %>%
  distinct(nom_espece, taxon_group, habitat_breadth, specialization, confidence, in_france) %>%
  filter(in_france == 1, confidence == 3) %>%
  arrange(desc(specialization)) %>%
  head(8)
```

# Defining habitat for *Apodemus sylvaticus*

## Scores of the Wood mouse

We filter the dictionary for *Apodemus sylvaticus* and join the EUNIS labels
and official colors from `ref_eunis`.

```{r apsy_dict}
ref_l2 <- ref_eunis %>% filter(level == "L2") %>% select(code, nomenclature, couleur)

apsy <- eunis_species_dict %>%
  filter(nom_espece == "Apodemus_sylvaticus") %>%
  left_join(ref_l2, by = c("code_eunis" = "code"))

apsy %>%
  distinct(habitat_breadth, specialization, confidence, in_france)
```

The Wood mouse is a generalist: 25 of the 45 classes are at least secondary
habitat (`habitat_breadth`), and its specialization index is close to 0.

```{r apsy_profile, fig.height=8}
apsy_long <- apsy %>%
  select(code_eunis, nomenclature, suitability, weight_breeding, weight_foraging, permeability) %>%
  tidyr::pivot_longer(c(suitability, weight_breeding, weight_foraging, permeability),
                      names_to = "score", values_to = "value") %>%
  mutate(label = factor(paste(code_eunis, substr(nomenclature, 1, 45)),
                        levels = rev(paste(apsy$code_eunis, substr(apsy$nomenclature, 1, 45)))),
         score = factor(score, levels = c("suitability", "weight_breeding", "weight_foraging", "permeability")))

ggplot(apsy_long, aes(value, label)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  facet_wrap(~score, nrow = 1) +
  scale_x_continuous(breaks = c(0, 5, 10)) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7)) +
  labs(x = "Score (0–10)", y = NULL, title = "Apodemus sylvaticus – scores per EUNIS Level 2 class")
```

## Understanding habitat values

As in the OCS-GE guide, the suitability score can be read as habitat quality
classes:

- Non-habitat (suitability = 0):

```{r ex_nonhabitat}
apsy %>% filter(suitability == 0) %>% pull(nomenclature)
```

- Very poor habitat (0 < suitability ≤ 3):

```{r ex_verypoor}
apsy %>% filter(suitability > 0, suitability <= 3) %>% pull(nomenclature)
```

- Poor to secondary habitat (3 < suitability ≤ 7):

```{r ex_poor}
apsy %>% filter(suitability > 3, suitability <= 7) %>% pull(nomenclature)
```

- Good habitat (suitability > 7):

```{r ex_good}
apsy %>% filter(suitability > 7) %>% pull(nomenclature)
```

# Mapping

## The EUNIS land cover layer

`get_eunis_data()` downloads the EUNIS raster for any region of interest from
the EEA map service. To keep this guide reproducible offline, the Metaleurop
layer is shipped with the package; it was produced with:

```{r get_eunis, eval=FALSE}
data("roi_metaleurop")
eunis_lc <- get_eunis_data(roi_metaleurop, level = "L2", resolution = 100)
```

```{r load_eunis}
eunis_lc <- load_raster_extdata("eunis_metaleurop_L2.tif")
eunis_lc
```

The raster is categorical. Its level table gives the EUNIS code of each pixel
value. Note the combined code `"D5/F6"`: the EEA service draws D5 (sedge and
reedbeds) and F6 (garrigue) with the same color, so these pixels cannot be
told apart (see `?get_eunis_data`).

```{r eunis_levels}
lv <- terra::levels(eunis_lc)[[1]]
names(lv) <- c("value", "code")
lv %>% left_join(ref_l2, by = "code")
```

## From species scores to rasters

We write a small helper that replaces each EUNIS class by a species score. For
combined codes such as `"D5/F6"`, the scores of the two classes are averaged.

```{r helper}
eunis_species_raster <- function(lc, species, var = "suitability",
                                 dict = eunis_species_dict) {
  lv <- terra::levels(lc)[[1]]
  names(lv) <- c("value", "code")
  d <- dict[dict$nom_espece == species, ]
  if (nrow(d) == 0) stop("Species not found: ", species)
  val <- vapply(strsplit(lv$code, "/"),
                function(cc) mean(d[[var]][match(cc, d$code_eunis)]), numeric(1))
  r <- lc
  levels(r) <- NULL                       # drop categories, keep pixel ids
  out <- terra::classify(r, cbind(lv$value, val))
  names(out) <- var
  out
}
```

```{r apsy_rasters}
apsy_suit <- eunis_species_raster(eunis_lc, "Apodemus_sylvaticus", "suitability")
apsy_breed <- eunis_species_raster(eunis_lc, "Apodemus_sylvaticus", "weight_breeding")
apsy_forag <- eunis_species_raster(eunis_lc, "Apodemus_sylvaticus", "weight_foraging")
apsy_res <- eunis_species_raster(eunis_lc, "Apodemus_sylvaticus", "resistance")
```

## Static maps

```{r apsy_staticmap, fig.height=9}
eunis_lc_lab <- eunis_lc
levels(eunis_lc_lab) <- lv %>%
  left_join(ref_l2, by = "code") %>%
  transmute(value, eunis = ifelse(is.na(nomenclature), code, paste(code, nomenclature)))
eunis_cols <- lv %>%
  left_join(ref_l2, by = "code") %>%
  transmute(eunis = ifelse(is.na(nomenclature), code, paste(code, nomenclature)),
            couleur = ifelse(is.na(couleur), "#BDBDBD", couleur))

p0 <- ggplot() +
  geom_spatraster(data = eunis_lc_lab) +
  scale_fill_manual(values = setNames(eunis_cols$couleur, eunis_cols$eunis),
                    na.value = "transparent", name = "EUNIS L2") +
  theme_minimal() +
  theme(legend.text = element_text(size = 7)) +
  labs(title = "Land cover (EUNIS Level 2)")

p1 <- ggplot() +
  geom_spatraster(data = apsy_suit) +
  scale_fill_gradient(low = "#ffffe5", high = "#004529", limits = c(0, 10),
                      na.value = "transparent", name = "Suitability") +
  theme_minimal() + labs(title = "Suitability")

p2 <- ggplot() +
  geom_spatraster(data = apsy_breed) +
  scale_fill_gradient(low = "#fcfbfd", high = "#3f007d", limits = c(0, 10),
                      na.value = "transparent", name = "Breeding") +
  theme_minimal() + labs(title = "Breeding weight")

p3 <- ggplot() +
  geom_spatraster(data = apsy_forag) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b", limits = c(0, 10),
                      na.value = "transparent", name = "Foraging") +
  theme_minimal() + labs(title = "Foraging weight")

p4 <- ggplot() +
  geom_spatraster(data = apsy_res) +
  scale_fill_gradient(low = "#fff5f0", high = "#67000d", trans = "log10", limits = c(1, 100),
                      na.value = "transparent", name = "Resistance") +
  theme_minimal() + labs(title = "Resistance (1–100, log scale)")

p0
(p1 | p2) / (p3 | p4) +
  plot_annotation(title = "Apodemus sylvaticus – habitat layers from EUNIS")
```

For the Wood mouse, the arable land (I1) that dominates the Metaleurop area is
good habitat (8/10), as are the woodland (G1) and gardens and parks (I2).
Built-up areas (J1, J2) are secondary habitat (5/10). Roads (J4) are nearly
unusable (1/10) and form the strongest barriers (resistance 63).

## Interactive synthesis

```{r leaflet_apsy, fig.width=9, fig.height=6}
to_wgs <- function(r) terra::project(r, "EPSG:4326", method = "near")

pal_suit <- colorNumeric("YlGn", domain = c(0, 10), na.color = "transparent")
pal_breed <- colorNumeric("Purples", domain = c(0, 10), na.color = "transparent")
pal_forag <- colorNumeric("Blues", domain = c(0, 10), na.color = "transparent")
pal_res <- colorNumeric("Reds", domain = c(0, 2), na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(to_wgs(apsy_suit), colors = pal_suit, opacity = 0.7, group = "Suitability") %>%
  addRasterImage(to_wgs(apsy_breed), colors = pal_breed, opacity = 0.7, group = "Breeding") %>%
  addRasterImage(to_wgs(apsy_forag), colors = pal_forag, opacity = 0.7, group = "Foraging") %>%
  addRasterImage(log10(to_wgs(apsy_res)), colors = pal_res, opacity = 0.7, group = "Resistance (log10)") %>%
  addLayersControl(
    baseGroups = c("Suitability", "Breeding", "Foraging", "Resistance (log10)"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(pal = pal_suit, values = c(0, 10), title = "Score (0–10)", position = "bottomright")
```

# Multi-species comparison

The same helper works for any species. Below are the species used in the
OCS-GE guide. Suitability shows where each one can live, and resistance
shows what blocks its movement.

```{r multi_species, fig.height=8}
species <- c("Apodemus_sylvaticus", "Turdus_merula", "Passer_domesticus",
             "Sorex_araneus", "Crocidura_russula", "Microtus_arvalis")

suit_stack <- rast(lapply(species, function(s) eunis_species_raster(eunis_lc, s, "suitability")))
names(suit_stack) <- gsub("_", " ", species)

ggplot() +
  geom_spatraster(data = suit_stack) +
  facet_wrap(~lyr, ncol = 3) +
  scale_fill_gradient(low = "#ffffe5", high = "#004529", limits = c(0, 10),
                      na.value = "transparent", name = "Suitability") +
  theme_minimal() +
  theme(axis.text = element_blank(), strip.text = element_text(face = "italic")) +
  labs(title = "Suitability from EUNIS for six species")
```

```{r multi_species_res, fig.height=8}
res_stack <- rast(lapply(species, function(s) eunis_species_raster(eunis_lc, s, "resistance")))
names(res_stack) <- gsub("_", " ", species)

ggplot() +
  geom_spatraster(data = res_stack) +
  facet_wrap(~lyr, ncol = 3) +
  scale_fill_gradient(low = "#fff5f0", high = "#67000d", trans = "log10", limits = c(1, 100),
                      na.value = "transparent", name = "Resistance") +
  theme_minimal() +
  theme(axis.text = element_blank(), strip.text = element_text(face = "italic")) +
  labs(title = "Resistance from EUNIS for six species")
```

The two birds never exceed a resistance of about 16 because they fly, whereas
roads reach 63 for all four small mammals. Built-up areas (J1) are core
habitat for the house sparrow and the blackbird (10/10), secondary for the
Wood mouse (5/10) and unsuitable for the common vole (0/10). The common vole
follows grassland (E2) and arable land (I1), while the common shrew prefers
woodland (G1) and avoids arable land (3/10).

A compact way to compare the species is to plot their scores for the classes
present in the study area:

```{r multi_species_profile, fig.height=5}
present <- unlist(strsplit(lv$code, "/"))

eunis_species_dict %>%
  filter(nom_espece %in% species, code_eunis %in% present) %>%
  left_join(ref_l2, by = c("code_eunis" = "code")) %>%
  mutate(nom_espece = gsub("_", " ", nom_espece),
         label = paste(code_eunis, substr(nomenclature, 1, 30))) %>%
  ggplot(aes(nom_espece, label, fill = suitability)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = suitability), size = 3) +
  scale_fill_gradient(low = "#ffffe5", high = "#004529", limits = c(0, 10)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, face = "italic")) +
  labs(x = NULL, y = NULL, fill = "Suitability",
       title = "Suitability of the EUNIS classes present at Metaleurop")
```

# Toward connectivity

The suitability raster is the **source** layer and the resistance raster is
the **resistance** layer for `compute_dispersal(method = "omniscape")`. Unlike the
OCS-GE dictionary, `resistance` is already on the 1–100 log scale, so no
exponential transformation is needed.

To combine them with soil contamination maps, resample them onto the same grid
as the contamination raster with nearest-neighbour interpolation:

```{r to_grid}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
apsy_suit_grid <- terra::project(apsy_suit, ground_cd, method = "near")
apsy_res_grid <- terra::project(apsy_res, ground_cd, method = "near")
c(apsy_suit_grid, apsy_res_grid)
```

```{r omniscape_eunis, eval=FALSE}
res_m <- 500                                   # dispersal radius in metres
radius_px <- res_m / terra::res(ground_cd)[1]
compute_dispersal(
  x = apsy_suit_grid,
  method = "omniscape",
  options = list(resistance = apsy_res_grid, radius = radius_px)
)
```

See [Landscape Connectivity (Omniscape)](core_Omniscape_Connectivity.html) for
running and interpreting the connectivity model.

Note that the EUNIS map is 100 m, so the landscape structure after
resampling is coarser than with OCS-GE. Use EUNIS when the study area is
outside France, or when the species depends on habitats that OCS-GE does not
separate (wetlands, grassland types, dunes).
