---
title: "Berisp-like: a trophic model of contamination"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of Trophic Contamination}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=10,
  fig.height=5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(spacemodR)
library(rstan)
library(dplyr)
library(terra)
```

```{r add_function, echo=FALSE}
# ADDITIONAL FUNCTION
read_sf_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(sf::read_sf(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
load_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(readRDS(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
```

# Define a Spacemodel

## Habitat

```{r}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
names_hab = c("soil", "plant", "earthworm", "carabid", "mamHerb", "mamInsect")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)

terra::plot(stack_habitat)
```

## Trophic web

```{r}
trophic_df <- trophic() |>
  add_link("soil", "plant", 1) |>
  add_link("soil", "earthworm", 1) |>
  add_link("soil", "carabid", 1) |>
  # mamHerb
  add_link("soil", "mamHerb", 2/100) |>
  add_link("plant", "mamHerb", 90/100) |>
  add_link("earthworm", "mamHerb", 4/100) |>
  add_link("carabid", "mamHerb", 4/100) |>
  # mamInsect
  add_link("soil", "mamInsect", 2/100) |>
  add_link("earthworm", "mamInsect", 49/100) |>
  add_link("carabid", "mamInsect", 49/100)
```

```{r}
plot(trophic_df)
```

## create the spacemodel

```{r}
spcmdl_trophic_fixed <- spacemodel(stack_habitat, trophic_df)
```


# Contaminantion by Cadmium

```{r load_data, echo=FALSE}
fit_simple_veg_cd <- load_safe("raw_data/fit_simple_veg_cd.rds")
fit_simple_worm_cd <- load_safe("raw_data/fit_simple_worm_cd.rds")
```

## Vegetation

```{r, echo=FALSE}
fit_pars <- rstan::extract(fit_simple_veg_cd, c("beta0", "beta1"))
df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1")
))
inter_veg = median(df_pars$beta0, na.rm=TRUE)
slope_veg = median(df_pars$beta1, na.rm=TRUE)
```


## Earthworm

For Cadmium, the equation is given by (Ma et al., 2004):

$$
\log C_{earthworm} = a + b \times \log C_{soil} + c \times \log OM + d \times pH
$$

```{r, echo=FALSE}
fit_pars <- rstan::extract(fit_simple_veg_cd, c("beta0", "beta1"))
df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1")
))
inter_worm = median(df_pars$beta0, na.rm=TRUE)
slope_worm = median(df_pars$beta1, na.rm=TRUE)
```

## Carabid

```{r}
inter_carab = -1
slope_carab = 0.6
```


## Transfer of food and contaminant

$$
C_{consumer} = \frac{b_{resource}}{b_{consumer}} \times  C_{resource} \times \frac{k_{up}}{k_{out}} \left( 1 - \exp^{- c_{out} \times a} \right)
$$
where:

- $b_{resource}$: average individual/item biomass of resource $[g]$,
- $b_{consumer}$: average individual/item biomass of consumer $[g]$,
- $C_{resource}$: concentration of substance in resource (e.g. $ppm=[ug.g^{-1}]=[mg.kg^{-1}]$),
- $C_{resource}$: concentration of substance in resource (e.g. $ppm=[ug.g^{-1}]=[mg.kg^{-1}]$),
- $k_{up}$: assimilation efficiency of food  $[n.d.]$,
- $k_{out}$: excretion rate of food, $[day^{-1}]$
- $a$: average age of the consumer $[day]$,

### Average fresh biomass of items and indiviuals

w  weight of species [kg]
(1.0  102
for S. araneus, 3.0  102
for M. agrestis,
and 2.1  102
for C. glareolus) [11,12,15]

```{r}
# b_apsy = 
b_vole = 50
b_shrew = 10
```

### Uptake and Excretion rate

- Uptake: The uptake rate constant of metals from food (kx,n,in) depends
on the food ingestion rate constant (kn
) and the metal assimilation efficiency from the food matrix (px,a
) (Eqn. 3). The
assimilation efficiency of cadmium in small mammals generally is low and depends on external conditions, such as pH
in the gastrointestinal tract, available biological ligands in
food, and stability of these metal complexes in the food matrix
[21]. At present, the mechanisms of metal uptake via ingestion
are not fully understood [20,22]; therefore, metal assimilation
efficiency cannot be modeled mechanistically. Instead, an empirical cadmium assimilation efficiency (px,a
) of 0.51% (Supplemental Data; http://dx.doi.org/10.1897/2006-518.S1) was
used to predict the uptake rate constant of cadmium from food
(kx,n,in) (Eqn. 3). This empirical cadmium assimilation efficiency is a geometric mean value of data from 10 long-term
experimental studies, including various experimental designs.
kx,n,in n  k ·px,a (3)
where
kn  food ingestion rate constant
[kg food/kg body wet wt/d]
px,a  assimilation efficiency of cadmium from food
(empirical value, see Supplemental Data; http://dx.
doi.org/10.1897/2006-518.S1) (0.51) [%]

```{r, eval=FALSE}
# k_up_vole
# k_ex_vole
```


### Transfer equation

```{r, eval=FALSE}
# /!\ Faudrait changer le code pour pouvoir faire un truc de ce tyle :
direct_intakes <- intake(spcmdl_trophic_fixed) |>
  add_intake("soil", "plant", ~ inter_veg + slope_veg*x) |>
  add_intake("soil", "earthworm", ~ inter_veg + slope_veg*x) |>
  add_intake("soil", "carabid", ~ inter_veg + slope_veg*x) |>
  default = NA, # for all other default is NA = no transfer or 1 full transfer ?
)
```

```{r, eval=FALSE}

# ATTENTION DANS LES FONCTION, ON AIME TRAVAILLER EN LOG !!!
direct_intakes <- intake(spcmdl_trophic_fixed,
  "soil -> plant"       = ~ inter_veg + slope_veg*x,  
  "soil -> earthworm"   = ~ inter_worm + slope_worm*x, 
  "soil -> carabid"     = ~ inter_carab + slope_carab*x, 
   # mamHerb
  "soil -> mamHerb"     = ~ b_soil_mamHerb / b_mamHerb * /0.36,
  "plant -> mamHerb"    = ~ 10^x/0.77,
  "earthworm -> mamHerb"= ~ 10^x/0.77,
  "carabid -> mamHerb"  = ~ 10^x/0.77,
  # mamInsect
  "soil -> mamInsect"     = ~ 10^x/0.77,
  "plant -> mamInsect"    = ~ 10^x/0.77,
  "earthworm -> mamInsect"= ~ 10^x/0.77,
  "carabid -> mamInsect"  = ~ 10^x/0.77,
  default = 1, # for all other default is 1
  normalize = FALSE # TRUE would weight every link to sum at 1
)
```


## dispersion

```{r, eval=FALSE}
# no dispersal
fixed_kernels <- list(
  soil  = NA, plant = NA, earthworm = NA,
  carabid=NA,  mamHerb = NA, mamInsect = NA)
```


```{r, eval=FALSE}
spcmdl_trophic_fixed_risk <- transfer(
  spcmdl_trophic_fixed,
  fixed_kernels,
  direct_intakes,
  exposure_weighting="potential")
```

