---
title: "Habitat and Connectivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Habitat and Connectivity}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=10,
  fig.height=5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

In this vignette, we explore the database linking OCS-GE
(French GIS layer for Occupation du Sol à Grande Échelle) 
labels with species labels.

We will analyze how different land cover types act as habitats with
varying qualities or resistances for different species,
starting with the Wood mouse (Apodemus sylvaticus).

```{r setup}
library(spacemodR)
library(dplyr)
library(ggplot2)
library(terra)
library(tidyterra)
library(sf)
library(leaflet)
library(patchwork)
```

# Defining Habitat

## Weighted species habitat with OCS-GE layer for *Apodemus sylvaticus*

First, we load the necessary datasets provided by spacemodR:

1. The species dictionary assigning weights to land cover codes.
2. The OCS-GE nomenclature reference (containing names and specific hex colors).
3. The spatial layer for OCS-GE (here, we use the ocsge_metaleurop dataset as our base map).


```{r load_data}
# Load the species dict and reference tables
data("ocsge_species_dict")
data("ref_ocsge")

# Load spatial data
data("ocsge_metaleurop") 
sf_ocsge <- ocsge_metaleurop
```

We filter our species dictionary for *Apodemus sylvaticus* and join it with the 
reference nomenclature to link the OCS-GE codes to their descriptive labels and predefined colors.

```{r clean_data}
# Filter for Apodemus sylvaticus
dfhab_Apsy <- ocsge_species_dict[
  grepl("Apodemus_sylvaticus",ocsge_species_dict$nom_espece, ignore.case = TRUE), 
  ]

# Join ref_ocsge with the species dictionary. 
# ref_ocsge$code_cs_ (with spaces) matches ocsge_species_dict$code_cs
df_merged_Apsy <- ref_ocsge %>%
  dplyr::left_join(dfhab_Apsy, by = c("code_cs_" = "code_cs"))

# Join with the spatial sf object. Assuming sf_ocsge uses 'code_cs' (without spaces)
sf_Apsy <- sf_ocsge %>%
  dplyr::left_join(df_merged_Apsy, by = "code_cs")
```

### Understanding Habitat Values

By examining the weight_global (wg), we can classify the landscape into 
different habitat qualities for the species:

- Non-habitat area (wg = 0):

```{r ex_nonhabitat}
df_merged_Apsy %>% 
  filter(weight_global == 0) %>%
  pull(nomenclature) %>%
  unique()
```

- Very Poor habitat (0 < wg <= 3):
```{r ex_verypoorhabitat}
df_merged_Apsy %>% 
  filter(weight_global > 0, weight_global <= 3) %>% 
  pull(nomenclature) %>% 
  unique()
```

- Poor habitat (3 < wg <= 7):

```{r ex_poorhabitat}
df_merged_Apsy %>% 
  filter(weight_global > 7) %>% 
  pull(nomenclature) %>% 
  unique()
```

- Good habitat (wg > 7):

```{r ex_goodhabitat}
df_merged_Apsy %>% 
  filter(weight_global > 7) %>% 
  pull(nomenclature) %>% 
  unique()
```

### Static Mapping

We will now visualize the spatial distribution of these metrics. 
We produce four maps focusing on Global Weight, Foraging Weight, 
Landscape Resistance, and finally the standard OCS-GE Land Cover.

```{r mouse_staticmap, echo=FALSE}
# 4. Map: OCS-GE Nomenclature with official colors
# Create a named vector for the manual color scale
ocsge_colors <- setNames(ref_ocsge$couleur, ref_ocsge$nomenclature)
p0 <- ggplot(sf_Apsy) +
  geom_sf(aes(fill = nomenclature), color = NA) +
  scale_fill_manual(values = ocsge_colors, name = "OCS-GE") +
  theme_minimal() +
  labs(title = "Land Cover (OCS-GE Nomenclature)")


# 1. Map: Global Weight
p1 <- ggplot(sf_Apsy) +
  geom_sf(aes(fill = weight_global), color = NA) +
  scale_fill_gradient(low = "#ffffe5", high = "#004529", na.value = "transparent", name = "Global\nWeight") +
  theme_minimal() +
  labs(title = "Apodemus sylvaticus - Global Habitat Weight")

# 2. Map: Foraging Weight
p2 <- ggplot(sf_Apsy) +
  geom_sf(aes(fill = weight_foraging), color = NA) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b", na.value = "transparent", name = "Foraging\nWeight") +
  theme_minimal() +
  labs(title = "Apodemus sylvaticus - Foraging Weight")

# 3. Map: Resistance
p3 <- ggplot(sf_Apsy) +
  geom_sf(aes(fill = resistance), color = NA) +
  scale_fill_gradient(low = "#fff5f0", high = "#67000d", na.value = "transparent", name = "Resistance") +
  theme_minimal() +
  labs(title = "Apodemus sylvaticus - Movement Resistance")

# Print maps
p0
p1
p2
p3
```

### Interactive Synthesis

To better explore the local variations and overlay different habitat
parameters for *Apodemus sylvaticus*, we combine them into a single
interactive leaflet map. You can toggle the layers in the top right corner.

```{r leaflet-apsy, fig.width=9, fig.height=6, echo=FALSE, cache=TRUE}
# Transform to WGS84 for Leaflet compatibility
sf_Apsy_wgs84 <- st_transform(sf_Apsy, 4326)

# Define color palettes
pal_wg <- colorNumeric("YlGn", domain = sf_Apsy_wgs84$weight_global, na.color = "transparent")
pal_wf <- colorNumeric("Blues", domain = sf_Apsy_wgs84$weight_foraging, na.color = "transparent")
pal_res <- colorNumeric("Reds", domain = sf_Apsy_wgs84$resistance, na.color = "transparent")

# Mapping strict pour Leaflet basé sur ref_ocsge
pal_ocsge <- colorFactor(palette = ref_ocsge$couleur, levels = ref_ocsge$nomenclature)

leaflet(sf_Apsy_wgs84) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%

  # Layer: OCS-GE
  addPolygons(
    fillColor = ~pal_ocsge(nomenclature), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "OCS-GE", label = ~paste(nomenclature)
  ) %>%

  # Layer: Global Weight
  addPolygons(
    fillColor = ~pal_wg(weight_global), color = NA, fillOpacity = 0.7,
    group = "Global Weight", label = ~paste("Global Weight:", weight_global)
  ) %>%

  # Layer: Foraging
  addPolygons(
    fillColor = ~pal_wf(weight_foraging), color = NA, fillOpacity = 0.7,
    group = "Foraging Weight", label = ~paste("Foraging:", weight_foraging)
  ) %>%

  # Layer: Resistance
  addPolygons(
    fillColor = ~pal_res(resistance), color = NA, fillOpacity = 0.7,
    group = "Resistance", label = ~paste("Resistance:", resistance)
  ) %>%

  # Controls
  addLayersControl(
    baseGroups = c("OCS-GE", "Global Weight", "Foraging Weight", "Resistance"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(pal = pal_ocsge, values = ref_ocsge$nomenclature, title = "OCS-GE", group = "OCS-GE", position = "bottomright")
```

## Multi-species Interactive Synthesis

Different species interact with the landscape in radically different ways. 
Below, we create a function to automatically generate the interactive 
Leaflet synthesis for any given species, allowing us to rapidly compare their habitat preferences.
### Turdus merula (Blackbird)

```{r plt_birdturdus}
turdus_hab <- join_ocsge_species(sf_ocsge, "Turdus_merula")
plot_species_habitat(turdus_hab)
```

### Passer domesticus (House Sparrow)

```{r plt_birdpasser}
passer_hab <- join_ocsge_species(sf_ocsge, "Passer_domesticus")
plot_species_habitat(passer_hab)
```

### Small Mammals (Sorex, Crocidura, Microtus arvalis)

```{r plt_sorex}
sorex_hab <- join_ocsge_species(sf_ocsge, "Sorex")
plot_species_habitat(sorex_hab)
```

```{r plt_crocidura}
crocidura_hab <- join_ocsge_species(sf_ocsge, "Crocidura")
plot_species_habitat(crocidura_hab)
```

```{r plt_microtus}
microtus_hab <- join_ocsge_species(sf_ocsge, "Microtus_arvalis")
plot_species_habitat(microtus_hab)
```

# Connectivity

## From habitat to raster of source and resistance

1. first step is to collect habitat of the OCS-GE as we did previously

```{r plt_apsy}
apsy_hab <- join_ocsge_species(sf_ocsge, "Apodemus_sylvaticus")
plot_species_habitat(apsy_hab)
```

2. then, we have to create an `habitat` object to benefit from `spacemodR` functionalities

```{r habitat_apsy}
apsy_habitat <- habitat(apsy_hab, habitat=TRUE, weight=apsy_hab$weight_global)
apsy_resistance <- habitat(apsy_hab, habitat=TRUE, weight=apsy_hab$resistance)
```


3. converting them in raster

```{r, echo=FALSE}
rast_Apsy_habitat <- load_raster_extdata("rast_Apsy_habitat.tif")
rast_Apsy_resistance_init <- load_raster_extdata("rast_Apsy_resistance_init.tif")
rast_Apsy_resistance <- load_raster_extdata("rast_Apsy_resistance.tif")
```

```{r}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
```


```{r plt_apsy_habitat, eval=FALSE}
rast_Apsy_habitat <- habitat_raster(ground_cd, apsy_habitat)
```

Note that for resistance, we have to use and exponential scale compare to the score in $[0-10]$
that with have in the data-set `ocsge_species_dict` to an exponential $[1-1000]$.

```{r plt_apsy_resistance}
rast_Apsy_resistance_init <- habitat_raster(ground_cd, apsy_resistance)
exp_scl <- function(x) { 10^( (log10(1000) / 10) * x )}
rast_Apsy_resistance <- exp_scl(rast_Apsy_resistance_init)
```

```{r, echo=FALSE}
# terra::writeRaster(rast_Apsy_habitat, "inst/extdata/rast_Apsy_habitat.tif", overwrite = TRUE)
# terra::writeRaster(rast_Apsy_resistance_init, "inst/extdata/rast_Apsy_resistance_init.tif", overwrite = TRUE)
# terra::writeRaster(rast_Apsy_resistance, "inst/extdata/rast_Apsy_resistance.tif", overwrite = TRUE)
```

```{r}
par(mfrow = c(1, 2))
plot(rast_Apsy_resistance_init)
plot(rast_Apsy_resistance)
par(mfrow = c(1, 1))
```


```{r, eval=FALSE}
terra::writeRaster(rast_Apsy_habitat, "julia_run/rast_Apsy_habitat.tif", overwrite = TRUE)
terra::writeRaster(rast_Apsy_resistance, "julia_run/rast_Apsy_resistance.tif", overwrite = TRUE)
```


## Kernel

## Omniscape

4. defining a radius

To define the radius, we have to know the real distance radius and then convert it into number of pixel.
Let say, we have a dispersal radius of 500m for *Apodemus sylvaticus*,
which is actually quite large for that species.

Here the map has pixels of 24m, so we use this large value for examples.

For pixel, the size is given by `res(raster)`, here it is a meter system:
 
```{r}
# get pixel size
terra::res(ground_cd)
# convert meter radius to pixel radius
m_radius = 500
px_radius = 500 / res(ground_cd)[1]
px_radius
```

```{bash, eval=FALSE}
docker run --rm \
  -v "$(pwd)/julia_run:/julia_run" \
  spacemodr Rscript /julia_run/run_dispersal.R \
  /julia_run/rast_Apsy_habitat.tif \
  /julia_run/rast_Apsy_resistance.tif \
  20.05 \
  omniscape_apsy_500m
```

```{bash, eval=FALSE}
docker run --rm \
  -v "$(pwd)/julia_run:/julia_run" \
  spacemodr Rscript /julia_run/run_dispersal.R \
  /julia_run/rast_Apsy_habitat.tif \
  /julia_run/rast_Apsy_resistance.tif \
  8.02 \
  omniscape_apsy_200m
```

```{r load_omniscape, echo=FALSE}
omniscape_apsy_200m <- load_raster_extdata("omniscape_apsy_200m.tif")
omniscape_apsy_500m <- load_raster_extdata("omniscape_apsy_500m.tif")
```


```{r plt_omniscape}
p_src = ggplot() +
  theme_minimal() +
  geom_spatraster(data = rast_Apsy_habitat) +
  scale_fill_gradient(low = "#ffffe5", high = "#004529", na.value = "transparent", name = "Global\nWeight") +
  labs(title = "Source")

p_res = ggplot() +
  theme_minimal() +
  geom_spatraster(data = rast_Apsy_resistance) +
  scale_fill_gradient(
    low = "#fff5f0", high = "#67000d", transform="log10",
    na.value = "transparent", name = "Resistance")+
  labs(title = "Resistance")

max_r200m <- global(omniscape_apsy_200m, fun = "max", na.rm = TRUE)
norm_r200m <- omniscape_apsy_200m / as.numeric(max_r200m)
p_omn_200m = ggplot() +
  theme_minimal() +
  geom_spatraster(data = norm_r200m) +
  scale_fill_gradientn(
    colours = c("white", "#11dddd", "#dddd11", "red"),
    trans = "sqrt", 
    na.value = "transparent") +
  labs(title = "Omniscape Connectivity (r=200m)", fill = "Probability")

max_r500m <- global(omniscape_apsy_500m, fun = "max", na.rm = TRUE)
norm_r500m <- omniscape_apsy_500m / as.numeric(max_r500m)
p_omn_500m = ggplot() +
  theme_minimal() +
  geom_spatraster(data = norm_r500m) +
  scale_fill_gradientn(
    colours = c("white", "#11dddd", "#dddd11", "red"),
    trans = "sqrt", na.value = "transparent") +
  labs(title = "Omniscape Connectivity (r=500m)", fill = "Probability")

gridExtra::grid.arrange(p_src, p_res, p_omn_200m, p_omn_500m, ncol=2)
```


