---
title: "Inference Direct"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Inference_Transfer}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=10,
  fig.height=5
)
```

```{r setup}
library(spacemodR)
# Additional package for further analysis
library(rstan)
library(ggplot2)
library(dplyr)
library(purrr)
library(sf)
library(terra)
library(tidyterra)
library(scales)
```

```{r echo=FALSE}
# ADDITIONAL FUNCTION
read_sf_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(sf::read_sf(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
load_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(readRDS(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}

```


```{r}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
sf_metaleurop_2010 <- read_sf_safe("raw_data/sf_metaleurop_2006_2010.gpkg")
sf_metaleurop_2025 <- read_sf_safe("raw_data/sf_metaleurop_2025.gpkg")
```

```{r}
data(sf_micromammals)

# ADD LOG_10 COLUMNS
sf_train <- sf_micromammals |>
  dplyr::mutate(
    log10_cd_S = log10(cd_S),
    log10_cd_WB_FW = log10(cd_WB_FW))

# ADD GROUP HERBIVOE vs INSECTIVORE
lookup_table = data.frame(
    group = c('shrew', 'mouse', 'vole'),
    food = c('insectivore', 'herbivore', 'herbivore'),
    food_cat_num = c(2,1,1))
sf_train = dplyr::left_join(sf_train, lookup_table, by='group')
```

# Direct fix-point model of Soil-Organism

## Inference

```{r direct_stanData}
stan_data = list(
  N = nrow(sf_train),
  K = 2,
  x = sf_train$log10_cd_S,
  y = sf_train$log10_cd_WB_FW,
  x_cat = sf_train$food_cat_num,
  M = 100,
  x_sim = seq(min(sf_train$log10_cd_S), max(sf_train$log10_cd_S), length.out=100)
)
```


```{r direct_loadfit}
# fit_direct <- calibrate_direct(stan_data, chains=4, warmup=500, iter=1000)
# saveRDS(fit_direct, file="raw_data/fit_direct_cd.rds")
fit_direct <- load_safe("raw_data/fit_direct_cd.rds")
```

### Equation parameters

The equation of the direct model is given by:

$$
\log_{10}\left( C_{internal} \right) = \beta_0 + \beta_1  \log_{10}\left( C_{soil} \right) + beta_{cat} \log_{10}\left( C_{soil} \right) 
$$

```{r direct_loadparsfit}
fit_pars <- rstan::extract(fit_direct, c("beta0", "beta1", "beta_cat"))

df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1"),
  data.frame(value=fit_pars$beta_cat[,1], par="beta_herbivore"),
  data.frame(value=fit_pars$beta_cat[,2], par="beta_insectivore")
))
ggplot(data=df_pars) +
  theme_minimal() +
  scale_x_log10() +
  scale_fill_manual(
    name="parameters",
    values=c("#22aa33", "#aa2233","#3322aa", "#2233aa")) +
  geom_density(aes(value, fill=par), alpha=0.5) 
```

### Observation vs Prediction

```{r}
arr_sim <- rstan::extract(fit_direct, "y_sim")[[1]]
```

```{r}
categories <- c("herbivore", "insectivore")
quants <- c(0.025, 0.5, 0.975) # Modifie avec les quantiles que tu souhaites
df_sim <- map_dfr(1:dim(arr_sim)[3], function(i) {
  mat <- arr_sim[, , i]
  q_mat <- apply(mat, 2, quantile, probs = quants)
  q_mat %>%
    t() %>%
    as.data.frame() %>%
    mutate(
      var_id = 1:n(),
      x_sim = stan_data$x_sim,
      categorie = categories[i]
    )
})
```

```{r}
plt_sim <- ggplot() +
  theme_minimal() +
  scale_color_manual(values = c("#22aa33", "#aa2233")) +
  scale_fill_manual(values = c("#22aa33", "#aa2233")) +
  labs(
    x = "Soil Concentration (Cd) [log10 µg/g]",
    y = "Whole Body Concentration (Cd) [log10 µg/g]",
    color = "Food / Categorie", # Unifie les légendes
    fill = "Food / Categorie"
  ) +
  geom_ribbon(
    data = df_sim,
    aes(x = x_sim, ymin = `2.5%`, ymax = `97.5%`, fill = categorie), 
    alpha = 0.25 
  ) +
  geom_line(
    data = df_sim,
    aes(x = x_sim, y = `50%`, color = categorie),
    linewidth = 1
  )

plt_train <- plt_sim +
  geom_point(
    data=sf_train,
    aes(x=log10_cd_S, y=log10_cd_WB_FW, color=food)
  )
plt_train
```

## Predictive Test

The Cd concentration in body, dry weight, µg.g-1, is calculated from dat*a measured using equation given in Veltman et al. 2007 in Environmental Toxicology and Chemistry:

- voles: $\frac{1}{0.8} \left( 0.054 \times \text{Cd}_\text{liver} + 0.013 \times \text{Cd}_\text{kidney} \right)$
- shrews: $\frac{1}{0.8} \left( 0.07 \times \text{Cd}_\text{liver} + 0.019 \times\text{Cd}_\text{kidney} \right)$

Then, the Cd concentration in body, fresh weight, µg.g-1, is calculated 
from data measured using the Dry/fresh ratio:  $Wfresh = \frac{1}{4} Wdry $

```{r}
# ADD GROUP HERBIVORE vs INSECTIVORE
lookup_table = data.frame(
    Species = c('APSY', 'CRRU', 'SOMI'),
    group = c('vole', 'shrew', 'shrew'),
    coef_WB_CD_liver = c(0.054, 0.07, 0.07),
    coef_WB_CD_kidney = c(0.013, 0.019, 0.019),
    food = c('herbivore', 'insectivore', 'insectivore'),
    food_cat_num = c(2,1,1))

sf_test = dplyr::left_join(sf_metaleurop_2025, lookup_table, by='Species') |>
  dplyr::mutate(cd_WB_DW = 1/0.8*(coef_WB_CD_liver*CdInLiver + coef_WB_CD_kidney*CdInKidneys)) |>
  dplyr::mutate(cd_WB_FW = cd_WB_DW*1/4) |>
  dplyr::mutate(log10_cd_WB_FW = log10(cd_WB_FW))
```

Collect corresponding soil value

```{r}
sf_test <- st_transform(sf_test, crs(ground_cd))

centroids <- st_centroid(sf_test)
val_centroid <- terra::extract(ground_cd, centroids)
val_mean_trapline <- terra::extract(ground_cd, vect(sf_test), fun = mean, na.rm = TRUE)

sf_test$log10_cd_S_centroid <- val_centroid[, 2]
sf_test$log10_cd_S_meanLine <- val_mean_trapline[, 2]
```

Big points are 

```{r}
sf_full <- data.frame(
  food =           c(sf_train$food, sf_test$food),
  set =            c(rep("train", nrow(sf_train)), rep("test", nrow(sf_test))),
  log10_cd_S =     c(sf_train$log10_cd_S, sf_test$log10_cd_S_meanLine),
  log10_cd_WB_FW = c(sf_train$log10_cd_WB_FW, sf_test$log10_cd_WB_FW)
)
sf_full$set <- factor(sf_full$set, levels = c("train", "test"))

plt_full <- plt_sim +
  geom_point(
    data = subset(sf_full, set == "train"),
    aes(x = log10_cd_S, y = log10_cd_WB_FW, color = food),
    shape = 16, alpha = 0.4, size = 1.5
  ) +
  geom_point(
    data = subset(sf_full, set == "test"),
    aes(x = log10_cd_S, y = log10_cd_WB_FW, fill = food),
    shape = 21, color = "black", stroke = 0.8, size = 2.5
  )
plt_full
```

## Mapping

Here we are going to compute the map of predicted concentration over the landscape.

Let first recapture the value of median:

```{r}
dp_q50 <- df_pars %>%
  dplyr::group_by(par) %>%
  dplyr::summarise(median_value = median(value, na.rm=TRUE))

ls_q50 <- setNames(as.list(dp_q50$median_value), dp_q50$par)
```

```{r}
names_hab = c("soil", "herbivore", "insectivore")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)
trophic_df <- trophic() |>
  add_link("soil", "herbivore") |>
  add_link("soil", "insectivore")
spcmdl_direct <- spacemodel(stack_habitat, trophic_df)

direct_kernels <- list(soil = NA, herbivore = NA, insectivore = NA)

direct_intakes <- intake(spcmdl_direct,
  "soil -> herbivore"       = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_herbivore*x,  
  "soil -> insectivore"     = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_insectivore*x,
  default = 1, # for all other default is 1
  normalize = FALSE # TRUE would weight every link to sum at 1
)

spcmdl_direct_risk <- transfer(
  spcmdl_direct,
  direct_kernels,
  direct_intakes,
  exposure_weighting="potential")
```

```{r}
r_risk = 10^spcmdl_direct_risk
ggplot() +
  geom_spatraster(data = r_risk) +
  facet_wrap(~lyr) +
  scale_fill_viridis_c(
    option = "magma",
    trans = "log10", # C'est ici que la magie opère
    labels = label_log(), # Affiche 10^1, 10^2, etc.
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Risk\n(log10-scaled)")
```

# Direct dist-buffer model of Soil-Organism

The idea here is to consider a buffer around the traplines. Let 100m first.

## Collect Buffer mean Value

```{r}
sf_train_b <- st_transform(sf_metaleurop_2010, crs(ground_cd))
sf_train_b100 <- st_buffer(sf_train_b, dist = 100)

centroids <- st_centroid(sf_train_b)
val_centroid <- terra::extract(ground_cd, centroids)
val_mean_trapline <- terra::extract(ground_cd, vect(sf_train_b), fun = mean, na.rm = TRUE)
val_mean_b100 <- terra::extract(ground_cd, vect(sf_train_b100), fun = mean,  na.rm = TRUE)

sf_train_b100$log10_cd_S_centroid <- val_centroid[, 2]
sf_train_b100$log10_cd_S_meanLine <- val_mean_trapline[, 2]
sf_train_b100$log10_cd_S_meanb100 <- val_mean_b100[, 2]
```

```{r}
melt_b100 <- sf_train_b100 %>%
  sf::st_drop_geometry() %>%
  tidyr::pivot_longer(
    cols=c("log10_cd_S_centroid", "log10_cd_S_meanLine", "log10_cd_S_meanb100",),
    values_to = "cd_S_collect", names_to = "method_collect")

ggplot(data=melt_b100) +
  theme_minimal() +
  scale_x_log10() +
  scale_color_viridis_d(option = "plasma") +
  geom_point(aes(cd_S, cd_S_collect, color=method_collect), alpha=0.5) +
  geom_abline(slope=1)
```

## Compute inference

```{r}
d00 = sf_train_b100 %>%
  dplyr::mutate(
    log10_cd_S = log10(cd_S),
    log10_cd_WB_FW = log10(cd_WB_FW),
  )
lookup_table = data.frame(
    group = c('shrew', 'mouse', 'vole'),
    food = c('insectivore', 'herbivore', 'herbivore'),
    food_cat_num = c(2,1,1))
d = dplyr::left_join(d00, lookup_table, by='group')

stan_data = list(
  N = nrow(d),
  K = 2,
  x = d$log10_cd_S_meanb100,
  y = d$log10_cd_WB_FW,
  x_cat = d$food_cat_num,
  M = 100,
  x_sim = seq(min(d$log10_cd_S), max(d$log10_cd_S), length.out=100)
)
```

```{r}
# fit_direct_b100 <- calibrate_direct(stan_data, chains=4, warmup=500, iter=1000)
# saveRDS(fit_direct_b100, file="raw_data/fit_direct_b100_cd.rds")
fit_direct_b100 <- load_safe("raw_data/fit_direct_b100_cd.rds")
```

```{r}
arr_sim_b100 <- rstan::extract(fit_direct_b100, "y_sim")[[1]]
```

```{r}
arr_sim <- arr_sim_b100
categories <- c("herbivore", "insectivore")
quants <- c(0.025, 0.5, 0.975) # Modifie avec les quantiles que tu souhaites
df_sim <- map_dfr(1:dim(arr_sim)[3], function(i) {
  mat <- arr_sim[, , i]
  q_mat <- apply(mat, 2, quantile, probs = quants)
  q_mat %>%
    t() %>%
    as.data.frame() %>%
    mutate(
      var_id = 1:n(),
      x_sim = stan_data$x_sim,
      categorie = categories[i]
    )
})
```

```{r}
plt_sim <- ggplot() +
  theme_minimal() +
  scale_color_manual(values = c("#22aa33", "#aa2233")) +
  scale_fill_manual(values = c("#22aa33", "#aa2233")) +
  labs(
    x = "Soil Concentration (Cd) [log10 µg/g]",
    y = "Whole Body Concentration (Cd) [log10 µg/g]",
    color = "Food / Categorie", # Unifie les légendes
    fill = "Food / Categorie"
  ) +
  geom_ribbon(
    data = df_sim,
    aes(x = x_sim, ymin = `2.5%`, ymax = `97.5%`, fill = categorie), 
    alpha = 0.25 
  ) +
  geom_line(
    data = df_sim,
    aes(x = x_sim, y = `50%`, color = categorie),
    linewidth = 1
  )

plt_train <- plt_sim +
  geom_point(
    data=sf_train,
    aes(x=log10_cd_S, y=log10_cd_WB_FW, color=food)
  )
plt_train
```

## Compare prediction with new data

```{r}
sf_full <- data.frame(
  food =           c(sf_train$food, sf_test$food),
  set =            c(rep("train", nrow(sf_train)), rep("test", nrow(sf_test))),
  log10_cd_S =     c(sf_train$log10_cd_S, sf_test$log10_cd_S_meanLine),
  log10_cd_WB_FW = c(sf_train$log10_cd_WB_FW, sf_test$log10_cd_WB_FW)
)
sf_full$set <- factor(sf_full$set, levels = c("train", "test"))

plt_full <- plt_sim +
  geom_point(
    data = subset(sf_full, set == "train"),
    aes(x = log10_cd_S, y = log10_cd_WB_FW, color = food),
    shape = 16, alpha = 0.4, size = 1.5
  ) +
  geom_point(
    data = subset(sf_full, set == "test"),
    aes(x = log10_cd_S, y = log10_cd_WB_FW, fill = food),
    shape = 21, color = "black", stroke = 0.8, size = 2.5
  )
plt_full
```

## Mapping

Here we are going to compute the map of predicted concentration over the landscape.

Let first recapture the value of median:

```{r}
fit_pars <- rstan::extract(fit_direct_b100, c("beta0", "beta1", "beta_cat"))

df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1"),
  data.frame(value=fit_pars$beta_cat[,1], par="beta_herbivore"),
  data.frame(value=fit_pars$beta_cat[,2], par="beta_insectivore")
))

dp_q50 <- df_pars %>%
  dplyr::group_by(par) %>%
  dplyr::summarise(median_value = median(value, na.rm=TRUE))

ls_q50 <- setNames(as.list(dp_q50$median_value), dp_q50$par)
```

```{r}
names_hab = c("soil", "herbivore", "insectivore")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)
trophic_df <- trophic() |>
  add_link("soil", "herbivore") |>
  add_link("soil", "insectivore")
spcmdl_direct <- spacemodel(stack_habitat, trophic_df)

direct_kernels <- list(soil = NA, herbivore = NA, insectivore = NA)

direct_intakes <- intake(spcmdl_direct,
  "soil -> herbivore"       = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_herbivore*x,  
  "soil -> insectivore"     = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_insectivore*x,
  default = 1, # for all other default is 1
  normalize = FALSE # TRUE would weight every link to sum at 1
)

spcmdl_direct_risk <- transfer(
  spcmdl_direct,
  direct_kernels,
  direct_intakes,
  exposure_weighting="potential")
```

```{r}
r_risk = 10^spcmdl_direct_risk
ggplot() +
  geom_spatraster(data = r_risk) +
  facet_wrap(~lyr) +
  scale_fill_viridis_c(
    option = "magma",
    trans = "log10", # C'est ici que la magie opère
    labels = label_log(), # Affiche 10^1, 10^2, etc.
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Risk\n(log10-scaled)")
```

# Direct kernel-buffer model of Soil-Organism



