---
title: "Berisp-like: a trophic model of contamination"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of Trophic Contamination}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=10,
  fig.height=5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(spacemodR)
library(rstan)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
library(terra)
```

```{r add_function, echo=FALSE}
# ADDITIONAL FUNCTION
read_sf_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(sf::read_sf(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
load_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(readRDS(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
```

# Define a Spacemodel

## Habitat

```{r}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
names_hab = c("soil", "plant", "earthworm", "carabid", "mamHerb", "mamInsect")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)
```

## Trophic web

```{r}
trophic_df <- trophic() |>
  add_link("soil", "plant", 1) |>
  add_link("soil", "earthworm", 1) |>
  add_link("soil", "carabid", 1) |>
  # mamHerb
  add_link("soil", "mamHerb", 2/100) |>
  add_link("plant", "mamHerb", 90/100) |>
  add_link("earthworm", "mamHerb", 4/100) |>
  add_link("carabid", "mamHerb", 4/100) |>
  # mamInsect
  add_link("soil", "mamInsect", 2/100) |>
  add_link("earthworm", "mamInsect", 49/100) |>
  add_link("carabid", "mamInsect", 49/100)
```

```{r}
plot(trophic_df)
```

## create the spacemodel

```{r}
spcmdl_trophic_fixed <- spacemodel(stack_habitat, trophic_df)
```


# Contaminantion by Cadmium

```{r load_data, echo=FALSE}
fit_simple_veg_cd <- load_safe("raw_data/fit_simple_veg_cd.rds")
fit_simple_worm_cd <- load_safe("raw_data/fit_simple_worm_cd.rds")
```

## Vegetation

```{r, echo=FALSE}
fit_pars <- rstan::extract(fit_simple_veg_cd, c("beta0", "beta1"))
df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1")
))
inter_veg = median(df_pars$beta0, na.rm=TRUE)
slope_veg = median(df_pars$beta1, na.rm=TRUE)
```


For Cadmium, the equation is given in Berisp:

$$
\log C_{vegetation} = 0.17 + 0.49 \times \log C_{soil} - 0.28 \times \log OM - 0.12 \times pH
$$

In Eco-SSL, authors use the following equation (Neperian logarithm):

$$
\ln(C_{plant}) = 0.546 * \ln(C_{soil}) - 0.475
$$

## Earthworm

For Cadmium, the equation is given by (Ma et al., 2004):

$$
\log C_{earthworm} = 2.92 + 0.747 \times \log C_{soil} - 0.5336 \times \log OM - 0.2101 \times pH
$$

```{r, echo=FALSE}
fit_pars <- rstan::extract(fit_simple_veg_cd, c("beta0", "beta1"))
df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1")
))
inter_worm = median(df_pars$beta0, na.rm=TRUE)
slope_worm = median(df_pars$beta1, na.rm=TRUE)
```

In Eco-SSL, the following equation is given (see Neperian logarithm):

$$
\ln(C_{earthworm}) = 0.795 * \ln(C_{soil}) + 2.114
$$

## Carabid

In Berisp, authors use the following equation:

$$
\log(C_{carabid}) = -1 + 0.6 * \log(C_{soil})
$$

```{r}
inter_carab = -1
slope_carab = 0.6
```


# Transfer of food and contaminant

$$
C_{consumer} = \frac{b_{resource}}{b_{consumer}} \times  C_{resource} \times \frac{k_{up}}{k_{out}} \left( 1 - \exp^{- c_{out} \times a} \right)
$$
where:

- $b_{resource}$: average individual/item biomass of resource $[g]$,
- $b_{consumer}$: average individual/item biomass of consumer $[g]$,
- $C_{resource}$: concentration of substance in resource (e.g. $ppm=[ug.g^{-1}]=[mg.kg^{-1}]$),
- $C_{resource}$: concentration of substance in resource (e.g. $ppm=[ug.g^{-1}]=[mg.kg^{-1}]$),
- $k_{up}$: assimilation efficiency of food  $[n.d.]$,
- $k_{out}$: excretion rate of food, $[day^{-1}]$
- $a$: average age of the consumer $[day]$,

## Diet and Body Mass

The diet and body mass of 5401 mammals and 9994 birds are in the attached datasets.

```{r}
data("DBFunc_MamFuncDat")
data("DBFunc_BirdFuncDat")
```

```{r, echo=FALSE}
library(stringr)
# 1. Définition de l'ordre trophique (du plus bas au plus élevé)
# Les noms correspondent aux catégories après la suppression du préfixe "Diet."
trophic_order <- c(
  "PlantO", # Tissus végétaux végétatifs (feuilles, tiges, racines) - Producteurs primaires
  "Fruit",  # Fruits / Baies
  "Seed",   # Graines / Noix
  "Nect",   # Nectar / Pollen
  "Inv",    # Invertébrés (Insectes, vers) - Consommateurs primaires/secondaires
  "Vfish",  # Poissons
  "Vect",   # Vertébrés ectothermes (reptiles, amphibiens)
  "Vend",   # Vertébrés endothermes (oiseaux, mammifères) - Super-prédateurs
  "Scav",   # Nécrophagie (charognards)
  "Vunk"    # Vertébrés inconnus / indéterminés
)
```


### Apodemus diet

In this dataset `DBFunc_MamFuncDat`, we can look at the estimated diet of all voles
species as well as their mean body mass.

```{r}
diet_apodemus = DBFunc_MamFuncDat[
  grepl("Apodemus", DBFunc_MamFuncDat$Scientific),
  ]

# transform data_set
diet_long <- diet_apodemus |>
  pivot_longer(
    cols = c(Diet.Inv, Diet.Vend, Diet.Vect, Diet.Vfish, Diet.Vunk, 
             Diet.Scav, Diet.Fruit, Diet.Nect, Diet.Seed, Diet.PlantO),
    names_to = "Diet_Category",
    values_to = "Value"
  ) |>
  mutate(
    Diet_Category = stringr::str_remove(Diet_Category, "Diet\\."),
    Diet_Category = factor(Diet_Category, levels = trophic_order)
  )

# create plot
ggplot(diet_long, aes(x = Diet_Category, y = Value, color = Scientific)) +
  geom_point(size = 3, alpha = 0.6, position = position_jitter(width = 0.15, height = 0)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Dietary Composition of Apodemus Species",
    x = "Diet Category",
    y = "Percentage / Value",
    color = "Species"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
  )
```

#### Concise Diet Descriptions (EltonTraits 1.0)

- `Diet.Inv`: % of invertebrates (insects, larves, worms, mollusks).
- `Diet.Vend`: % of endothermic vertebrates (birds and mammals).
- `Diet.Vect`: % of ectothermic vertebrates (reptiles and amphibians).
- `Diet.Vfish`: % of fish consumed.
- `Diet.Vunk`: % of unknown or unclassified vertebrates.
- `Diet.Scav`: % of scavenging activity (eating carrion/dead animals).
- `Diet.Fruit`: % of fruits and berries consumed.
- `Diet.Nect`: % of nectar, pollen, or plant exudates.
- `Diet.Seed`: % of seeds, nuts, grains, or cones.
- `Diet.PlantO`: % of other vegetative tissues (leaves, stems, roots, bark).


And also:

- `Diet.Source`: Reference/ID of the scientific data source.
- `Diet.Certainty`: Data certainty score (direct species study vs. genus-level extrapolation).

### Small mammals: voles and shrew

```{r}
diet_mammal <- DBFunc_MamFuncDat |>
  dplyr::filter(
    Scientific %in% 
      c("Microtus arvalis", "Myodes glareolus", "Apodemus sylvaticus", "Sorex araneus", "Crocidura russula")
    )

# transform data_set
diet_long <- diet_mammal |>
  pivot_longer(
    cols = c(Diet.Inv, Diet.Vend, Diet.Vect, Diet.Vfish, Diet.Vunk, 
             Diet.Scav, Diet.Fruit, Diet.Nect, Diet.Seed, Diet.PlantO),
    names_to = "Diet_Category",
    values_to = "Value"
  ) |>
  mutate(
    Diet_Category = stringr::str_remove(Diet_Category, "Diet\\."),
    Diet_Category = factor(Diet_Category, levels = trophic_order)
  )

# create plot
ggplot(diet_long, aes(x = Diet_Category, y = Value, color = Scientific)) +
  geom_point(size = 3, alpha = 0.6, position = position_jitter(width = 0.15, height = 0)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Dietary composition of some small-mammals species",
    x = "Diet Category",
    y = "Percentage / Value",
    color = "Species"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
  )
```

We can compared the results with what is used in the Berisp documentation:

| | grass | herbs | berries | seeds | earthworm | beetle | soil | 
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| bank vole (*Myodes glareolus*) | 20 | 20 | 26 | 24 | 4 | 4 | 2 |
| common vole (*Microtus arvalis*)| 42 | 41 | | 15 | | | 2 |
| wood mouse (*Apodemus sylvaticus*) | 6 | 6 | 12 | 58 | 8 | 8 | 2 |
| shrew (*Sorex araneus* or *Crocidura russula*) | | | | | 49 | 49 | 2 |


### Birds: little owl and black bird

```{r}
diet_bird <- DBFunc_BirdFuncDat |>
  dplyr::filter(
    Scientific %in% 
      c("Athene noctua", "Turdus merula")
    )

# transform data_set
diet_long <- diet_bird |>
  pivot_longer(
    cols = c(Diet.Inv, Diet.Vend, Diet.Vect, Diet.Vfish, Diet.Vunk, 
             Diet.Scav, Diet.Fruit, Diet.Nect, Diet.Seed, Diet.PlantO),
    names_to = "Diet_Category",
    values_to = "Value"
  ) |>
  mutate(
    Diet_Category = stringr::str_remove(Diet_Category, "Diet\\."),
    Diet_Category = factor(Diet_Category, levels = trophic_order)
  )

# create plot
ggplot(diet_long, aes(x = Diet_Category, y = Value, color = Scientific)) +
  geom_point(size = 3, alpha = 0.6, position = position_jitter(width = 0.15, height = 0)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Dietary composition of some small-mammals species",
    x = "Diet Category",
    y = "Percentage / Value",
    color = "Species"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
  )
```

## Energy Needs

We have the energy needs of 749 species, 97 mammals, 107 birds, 170 fishes,
51 reptiles, 11 amphibians, 110 crustacean, 65 arthropods, 75 protozoa

```{r}
data("FmrBT")
```

#### close species

The list is missing some species.

* Sorex and Crocidura (Shrews): the list lacks true shrews (Soricidae). When seeking a metabolic surrogate for temperate shrews (Sorex spp.) within this dataset, small temperate insectivorous bats, such as *Myotis lucifugus* and *Plecotus auritus*, serve as excellent physiological equivalents. Like true shrews, these bats are lightweight (often weighing between 5 and 12 grams) and operate under immense thermal pressure from temperate and boreal climates. Because they are strictly insectivorous, they share a highly active foraging strategy that demands a continuous supply of high-protein, easily digestible prey. Most importantly, the extreme energetic cost of flapping flight combined with their tiny body size forces these bats to run a hyper-metabolic "engine."

* *Myodes glareolus* (Bank Vole): the list is highly enriched with rodents that are phylogenetically very close to the bank vole. Most notably, *Cleithrionomys rutilus* belongs to the exact same genus (as Myodes and *Cleithrionomys* are synonymous in modern taxonomy). Additionally, species from the genera *Microtus* (like *M. agrestis* and *M. pennsylvanicus*) and *Arvicola* belong to the same family, Cricetidae, sharing identical microtine (vole-like) metabolic and toxicokinetic traits.

```{r}
energy_mammal <- FmrBT |>
  filter(
    SpeciesVerbatim %in% c(
      "Apodemus sylvaticus", "Plecotus auritus", "Myotis lucifugus",
      "Cleithrionomys rutilus",
      "Microtus agrestis", "Microtus pennsylvanicus")
  )

# plot
ggplot(data = energy_mammal, aes(x = Mass_g, y = FMR_kJ_d/Mass_g, color = SpeciesVerbatim)) +
  geom_point(size = 3.5, alpha = 0.8) + 
  labs(
    title = "Field Metabolic Rate (FMR) vs. Body Mass in Small Mammals",
    x = "Body Mass (g)",
    y = "Field Metabolic Rate (kJ/day/g)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )
```


* *Athene noctua* (Little Owl): true owls (Strigiformes) are completely absent
from this dataset. The closest relative available is *Phalaenoptilus nuttallii* (the common poorwill), which belongs to the *Caprimulgiformes* (nightjars). While it is a distinct family, nightjars share a nocturnal, insectivorous niche with small owls and belong to the same broader evolutionary landbird/Strisores lineage, making it the best available surrogate for nocturnal avian traits.

```{r}
energy_mammal <- FmrBT |>
  filter(SpeciesVerbatim %in% c(
      "Turdus merula", "Phalaenoptilus nuttallii")
  )

# plot
ggplot(data = energy_mammal, aes(x = Mass_g, y = FMR_kJ_d/Mass_g, color = SpeciesVerbatim)) +
  geom_point(size = 3.5, alpha = 0.8) + 
  labs(
    title = "Field Metabolic Rate (FMR) vs. Body Mass in Small Mammals",
    x = "Body Mass (g)",
    y = "Field Metabolic Rate (kJ/day/g)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

# Dose of Exposure

## Trophic transfer equation

```{r, eval=FALSE}
direct_fluxes <- flux(
    spcmdl_trophic_fixed,
    default = 1,    # for all other default is 1
    normalize=FALSE # TRUE would weight every link to sum at 1
  ) |>
  add_flux("soil", "plant", ~ inter_veg + slope_veg*x) |>
  add_flux("soil", "earthworm", ~ inter_veg + slope_veg*x) |>
  add_flux("soil", "carabid", ~ inter_veg + slope_veg*x)
```

## Population dispersion in landscape

```{r, eval=FALSE}
# no dispersal
fixed_kernels <- list(
  soil  = NA, plant = NA, earthworm = NA,
  carabid=NA,  mamHerb = NA, mamInsect = NA)
```


```{r, eval=FALSE}
spcmdl_trophic_fixed_risk <- transfer(
  spcmdl_trophic_fixed,
  fixed_kernels,
  direct_fluxes,
  exposure_weighting="potential")
```

## Computing Population Exposure

# Risk based on SSD
