---
title: "A simple example: a soil-target transfer with fix or moving target"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulation_soil_target}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.width = 12,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)

# Load libraries
library(spacemodR)
library(ggplot2)
library(sf)
library(terra)
library(tidyterra)

load_safe <- function(file_path) {
  paths <- c("", "../", "../../") |> paste0(file_path)
  valid_path <- paths[file.exists(paths)][1]
  if (!is.na(valid_path)) {
    return(readRDS(valid_path))
  }
  warning("File not found at: ", paste(paths, collapse = " OR "))
  return(NULL)
}
```

# Presentation of the Metaleurop Site and Data

The "Metaleurop" case involves a 40 km² site in northern France centered around
a former lead smelter. Soil levels around the plant are relatively high. 

Below are the extrapolated contamination maps for Cadmium (Cd), Lead (Pb), 
and Zinc (Zn) in the soil, derived from 595 sampling points
(the extrapolation methodology is not covered in this vignette).


```{r soil_contam}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
ground_pb <- load_raster_extdata("ground_concentration_pb_compressed.tif")
ground_zn <- load_raster_extdata("ground_concentration_zn_compressed.tif")
ground <- terra::rast(list(cd=ground_cd, pb=ground_pb, zn=ground_zn))

terra::plot(ground, main="Soil Concentration (log10)")
```


```{r soil_contam_plot}
ground_df <- as.data.frame(ground, xy = TRUE, na.rm = TRUE)

plt_init = ggplot(data = ground_df, aes(x = x, y = y)) +
  theme_minimal() +
  labs(title="Concentration (lo10 scaled)", x="longitude", y="latitude") +
  scale_fill_viridis_c()

plt_init + geom_raster(aes(fill=cd))

plt_init + geom_raster(aes(fill=pb))

plt_init + geom_raster(aes(fill=zn))
```


Since 2006, a total of 1426 micro-mammals (representing 9 different species) 
have been captured on-site to measure the bioaccumulation of metals in various
organs (kidneys, livers).

To understand the relationship between soil contamination and bioaccumulation, 
we can categorize these mammals into two trophic groups: herbivores and 
insectivores. By plotting the whole-body Cadmium concentration against
the soil Cadmium concentration at their capture locations, we can fit 
linear regression models to establish basic transfer equations.

```{r}
data("sf_micromammals")
trophic_cat =  ifelse(sf_micromammals$genus %in% c("crocidura", "sorex"), "insectivore", "herbivore")
sf_micromammals$trophic = factor(trophic_cat, levels = c("herbivore", "insectivore"))
ptsHerb = sf_micromammals[sf_micromammals$trophic=="herbivore", ]
ptsInsect = sf_micromammals[sf_micromammals$trophic=="insectivore", ]

lmH <- lm(log10(cd_WB_FW) ~ log10(cd_S), data = ptsHerb)
lmI <- lm(log10(cd_WB_FW) ~ log10(cd_S), data = ptsInsect)

ggplot(data = sf_micromammals) +
  theme_minimal() +
  labs(title="Cadmium (Cd): log10(Whole body) ~ log10(Soil)",
       x="Soil concentration (log10)",
       y="Whole body concentration (log10)") +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(
    name="Trophic category",
    values=c("#11aa44", "#aa4411")) +
  geom_point(aes(x=cd_S, y=cd_WB_FW, color=trophic)) +
  geom_abline(intercept=coef(lmH)[1], slope=coef(lmH)[2], color = "#11aa44", size=2) +
  geom_abline(intercept=coef(lmI)[1], slope=coef(lmI)[2], color = "#aa4411", size=1.5) +
  geom_abline(intercept= -1.2571, slope=0.4723, color = "red")
```

# Results of Fit

```{r}
fit_direct <- load_safe("raw_data/fit_direct_cd.rds")
fit_pars <- rstan::extract(fit_direct, c("beta0", "beta1", "beta_cat"))
df_pars <- dplyr::bind_rows(list(
  data.frame(value=fit_pars$beta0, par="beta0"),
  data.frame(value=fit_pars$beta1, par="beta1"),
  data.frame(value=fit_pars$beta_cat[,1], par="beta_herbivore"),
  data.frame(value=fit_pars$beta_cat[,2], par="beta_insectivore")
))
```


# Direct fix-point model of Soil-Organism

In this first approach, we will build a highly simplified spatial model.
We assume that individuals do not move (fixed points) and that contamination 
transfers directly from the soil to the target organisms (herbivorous and 
insectivorous mammals), bypassing intermediate trophic levels like plants 
or invertebrates.

## Step 1: Habitat Initialization

First, we need to define the spatial environment (habitat) for our compartments.
Since this is a simple direct-transfer model, the "habitat" for all entities 
is simply represented by the spatial grid of the soil contamination.

```{r}
# Define the compartments of our system
unique(sf_micromammals$genus)
names_hab = c("soil", "mamHerb", "mamInsect")

# Assign the Cadmium soil raster to all compartments as their base spatial extent
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)

terra::plot(stack_habitat)
```

## Step 2: Defining the Trophic Network

Next, we establish the links between our compartments. 
Using the `trophic()` function, we create direct pathways from the `soil` to 
both `mamHerb` and `mamInsect`.

```{r}
trophic_df <- trophic() |>
  add_link("soil", "mamHerb") |>
  add_link("soil", "mamInsect")

plot(trophic_df, shift=FALSE)
```

With the habitat stack and the trophic network defined, we can instantiate our base spatial model.

```{r}
spcmdl_simple <- spacemodel(stack_habitat, trophic_df)
```

## Step 3: Setting the Movement Kernels (Fixed Individuals)

Because this is a "fix-point" model, we assume the organisms 
stay exactly where they are. 
In spacemodR, we represent this lack of spatial 
dispersal by setting the movement kernels to `NA`.

```{r}
kernels <- list(
  soil  = NA,
  mamHerb = NA,
  mamInsect = NA
)
```

## Step 4: Parameterizing the Transfer Intakes

Now we define how much contamination transfers along the links we created. We use the equations derived from the linear regressions in the introduction.

For instance, the linear model for herbivores yielded roughly $y = 0.2x - 1$, meaning the intake formula can be written as `10^(-1 + 0.2*x)`.

```{r}
simple_intakes <- intake(spcmdl_simple,
  "soil -> mamHerb" = ~ 10^(-1 + 0.2*x),
  "soil -> mamInsect" = ~ 10^(-0.2 + 0.3*x),
  default = 1 # Default multiplier for any unassigned links
)

# Apply the kernels and intakes to compute the transfer through the model
spcmdl_transfer <- transfer(spcmdl_simple, kernels, simple_intakes)
```

## Step 5: Visualizing Model Predictions

The transfer function calculates the resulting concentration maps for our target mammals. We can visualize the predicted Cadmium levels for the herbivores, overlaid with the actual sampling locations

```{r}
df_raster <- as.data.frame(spcmdl_transfer, xy = TRUE, na.rm = TRUE)

ggplot() +
  geom_raster(data = df_raster, aes(x = x, y = y, fill = mamHerb)) +
  scale_fill_viridis_c() +
  geom_sf(data = sf_micromammals, color = "red", size = 1.5, alpha = 0.5) +
  theme_minimal() +
  labs(title = "Predicted Cadmium Concentration in Herbivores", 
       subtitle = "Red dots indicate mammal sampling locations",
       fill = "Predicted Cd (log10)")
```

## Step 6: Validation against Observations

To evaluate our model, we extract the predicted values at the exact 
spatial coordinates where the micro-mammals were captured and compare 
them directly to the observed whole-body concentrations.

```{r}
# Extract predicted values at capture points
predict_insect <- terra::extract(spcmdl_transfer[["mamInsect"]], ptsInsect)
ptsInsect$predict <- predict_insect[, "mamInsect"]

predict_herb <- terra::extract(spcmdl_transfer[["mamHerb"]], ptsHerb)
ptsHerb$predict <- predict_herb[, "mamHerb"]
```

If our model is perfectly accurate, the points should align closely with the 1:1 diagonal line (Predicted = Observed).

```{r}
ggplot() +
  theme_minimal() +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Model Validation: Predicted vs Observed", 
       x = "Observed Whole Body Concentration (log10)", 
       y = "Predicted Whole Body Concentration (log10)") + 
  geom_abline(slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(data = ptsInsect, aes(x = cd_WB_FW, y = predict), color = "#aa4411", alpha = 0.7) +
  geom_point(data = ptsHerb, aes(x = cd_WB_FW, y = predict), color = "#11aa44", alpha = 0.7) +
  annotate("text", x = 0.01, y = 10, label = "Insectivores", color = "#aa4411", hjust = 0) +
  annotate("text", x = 0.01, y = 5, label = "Herbivores", color = "#11aa44", hjust = 0)
```


```{r}
sf_predict = rbind(ptsInsect, ptsHerb)
ggplot(data = sf_predict) +
  theme_minimal() +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values=c("#11aa44", "#aa4411")) +
  geom_point(aes(x=cd_S, y=predict, color=trophic)) +
  geom_abline(intercept= -0.2, slope=0.3, color = "#aa4411" ) +
  geom_abline(intercept= -1,  slope=0.2, color = "#11aa44" ) +
  geom_abline(intercept= -1.2571, slope=0.4723, color = "red")
```


## Mapping

Here we are going to compute the map of predicted concentration over the landscape.

Let first recapture the value of median:

```{r}
dp_q50 <- df_pars %>%
  dplyr::group_by(par) %>%
  dplyr::summarise(median_value = median(value, na.rm=TRUE))

ls_q50 <- setNames(as.list(dp_q50$median_value), dp_q50$par)
```

```{r}
names_hab = c("soil", "herbivore", "insectivore")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)
trophic_df <- trophic() |>
  add_link("soil", "herbivore") |>
  add_link("soil", "insectivore")
spcmdl_direct <- spacemodel(stack_habitat, trophic_df)

direct_kernels <- list(soil = NA, herbivore = NA, insectivore = NA)

direct_intakes <- intake(spcmdl_direct,
  "soil -> herbivore"       = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_herbivore*x,  
  "soil -> insectivore"     = ~ ls_q50$beta0 + ls_q50$beta1*x + ls_q50$beta_insectivore*x,
  default = 1, # for all other default is 1
  normalize = FALSE # TRUE would weight every link to sum at 1
)

spcmdl_direct_risk <- transfer(
  spcmdl_direct,
  direct_kernels,
  direct_intakes,
  exposure_weighting="potential")
```

```{r direct_map, echo=FALSE}
r_risk = 10^spcmdl_direct_risk
map_directFix <- ggplot() +
  tidyterra::geom_spatraster(data = r_risk) +
  facet_wrap(~lyr) +
  scale_fill_viridis_c(
    option = "magma",
    trans = "log10",
    # labels = label_log(), # Affiche 10^1, 10^2, etc.
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Concentration Cd\n direct fix model\n(log10-scaled)")

map_directFix
```


# A simple Soil - Target model with Dispersal

```{r}
ground_cd <- load_raster_extdata("ground_concentration_cd_compressed.tif")
# names_hab = c("soil", "plant", "invert", "mamHerb", "mamInsect", "birdInsect")
names_hab = c("soil", "mamHerb", "mamInsect")
list_habitat <- lapply(names_hab, function(i) ground_cd)
stack_habitat <- raster_stack(list_habitat, names_hab)

terra::plot(stack_habitat)
```


```{r}
trophic_df <- trophic() |>
  #add_link("soil", "plant") |>
  #add_link("soil", "invert") |>
  add_link("soil", "mamHerb") |>
  add_link("soil", "mamInsect") # |>
  #add_link("soil", "birdInsect")

plot(trophic_df, shift=FALSE)
```


```{r}
spcmdl_simple <- spacemodel(stack_habitat, trophic_df)
```

```{r}
k_mamInsect <- compute_kernel(radius=100, GSD=25, size_std=1.5)
k_mamHerb <- compute_kernel(radius=500, GSD=25, size_std=1.5)

kernels <- list(
  soil  = NA,
  #plant = NA,
  #invert = NA,
  mamHerb = k_mamHerb,
  mamInsect = k_mamInsect#,
  #birdInsect = NA
)
```


```{r}
simple_intakes <- intake(spcmdl_simple,
  "soil -> mamHerb" = ~ 10^(-1 + 0.2*x),
  "soil -> mamInsect" = ~ 10^(-0.2 + 0.3*x),
  default = 1 # for all other default is 1
)
spcmdl_transfer <- transfer(spcmdl_simple, kernels, simple_intakes)
```

```{r}
terra::plot(spcmdl_transfer)
```

```{r}
df_raster <- as.data.frame(spcmdl_transfer, xy = TRUE, na.rm = TRUE)
ggplot() +
  geom_raster(data = df_raster, aes(x = x, y = y, fill = mamHerb)) +
  scale_fill_viridis_c() +
  geom_sf(data = sf_micromammals, color = "red", size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Ma carte Raster + Points", fill = "Valeur")
```


```{r}
ptsInsect = sf_micromammals[sf_micromammals$trophic=="insectivore", ]
ptsHerb = sf_micromammals[sf_micromammals$trophic=="herbivore", ]

predict_insect <- terra::extract(spcmdl_transfer[["mamInsect"]], ptsInsect)
ptsInsect$predict <- predict_insect[, "mamInsect"]

predict_herb <- terra::extract(spcmdl_transfer[["mamHerb"]], ptsHerb)
ptsHerb$predict <- predict_herb[, "mamHerb"]
```


```{r}
ggplot() +
  theme_minimal() +
  scale_x_log10() + scale_y_log10() +
  labs(title="Predicted ~  Observed", x = "Observed", y = "Predicted") + 
  geom_abline(slope=1) +
  geom_point(data = ptsInsect, color="#aa4411", 
             aes(x=cd_WB_FW, y=predict, color=trophic)) +
  geom_point(data = ptsHerb, color="#11aa44",
             aes(x=cd_WB_FW, y=predict, color=trophic))
```


```{r}
sf_predict = rbind(ptsInsect, ptsHerb)
ggplot(data = sf_predict) +
  theme_minimal() +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values=c("#11aa44", "#aa4411")) +
  geom_point(aes(x=cd_S, y=predict, color=trophic)) +
  geom_abline(intercept= -0.2, slope=0.3, color = "#aa4411" ) +
  geom_abline(intercept= -1,  slope=0.2, color = "#11aa44" ) +
  geom_abline(intercept= -1.2571, slope=0.4723, color = "red")
```


